---
title: "AR SuSiE-IW demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  run_analysis: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This page records the self-contained AR-correlation demonstration distributed
with the project code. It simulates data from the SuSiE-IW model and
compares in-sample and reference-LD SuSiE-RSS, collapsed-R, and AIW-N, with and
without concentration calibration.

The complete simulation and fits run when this page is rendered. Set the YAML
parameter `run_analysis` to `false` to skip the computationally nontrivial
analysis during a workflowr build.

## Analysis

```{r analysis, eval=params$run_analysis}
release_dir <- if (file.exists("01-collapsed-R.R")) {
  "."
} else {
  "code"
}
source(file.path(release_dir, "01-collapsed-R.R"))
source(file.path(release_dir, "02-AIW-and-AIW-N.R"))
source(file.path(release_dir, "03-SER-fallback-and-plot.R"))

# -------------------------------------------------------------------------
# Simulation settings
# -------------------------------------------------------------------------

seed <- 5L
J <- 500L
N <- 50000L
L <- 5L
heritability <- 1e-2
ar_correlation <- 0.9
true_lambda <- 1e-3
true_eta0 <- 200
causal <- c(100L, 400L)       # weak LD because the variants are far apart
effect_weights <- c(7, 4)
lambda_grid <- collapsed_r_default_lambda_grid()

collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

# -------------------------------------------------------------------------
# SuSiE-IW simulation: R | R0, followed by v | R, beta
# -------------------------------------------------------------------------

set.seed(seed)
R0 <- outer(seq_len(J), seq_len(J), function(j, k) {
  ar_correlation^abs(j - k)
})
Rbar_true <- (1 - true_lambda) * R0 + true_lambda * diag(J)

precision <- stats::rWishart(
  1L, df = true_eta0 + J + 1L,
  Sigma = solve(true_eta0 * Rbar_true)
)[, , 1L]
R_in <- stats::cov2cor(solve(precision))
R_in <- (R_in + t(R_in)) / 2
diag(R_in) <- 1

beta <- numeric(J)
beta[causal] <- effect_weights
beta <- beta * sqrt(
  heritability / as.numeric(crossprod(beta, R_in %*% beta))
)
noise <- as.numeric(t(chol(R_in)) %*% stats::rnorm(J)) / sqrt(N)
v <- as.numeric(R_in %*% beta) + noise

cat(sprintf(
  "Simulated J=%d, N=%d, h2=%g; causal LD=%.3f\n",
  J, N, heritability, R_in[causal[1L], causal[2L]]
))

# -------------------------------------------------------------------------
# Six fits. eta0_multiplier=1 gives the uncorrected method.
# -------------------------------------------------------------------------

susie_in <- fit_susie_rss(v, R_in, N, L = L)
susie_out <- fit_susie_rss(v, R0, N, L = L)

collapsed_raw <- susie_iw(
  v = v, R0 = R0, N = N, L = L,
  lambda_grid = lambda_grid,
  eta0_multiplier = 1,
  verbose = FALSE
)

collapsed_calibrated <- susie_iw(
  v = v, R0 = R0, N = N, L = L,
  lambda_grid = lambda_grid,
  eta0_multiplier = collapsed_eta0_multiplier,
  verbose = FALSE
)

aiwn_raw <- susie_aiw(
  v = v, R0 = R0, N = N, L = L, approximation = "N",
  lambda_grid = lambda_grid,
  eta0_multiplier = 1,
  verbose = FALSE
)
aiwn_calibrated <- susie_aiw(
  v = v, R0 = R0, N = N, L = L, approximation = "N",
  lambda_grid = lambda_grid,
  eta0_multiplier = aiwn_eta0_multiplier,
  verbose = FALSE
)

fits <- setNames(list(
  susie_in, susie_out, collapsed_raw, collapsed_calibrated,
  aiwn_raw, aiwn_calibrated
), c(
  "SuSiE-RSS: in-sample R",
  "SuSiE-RSS: out-of-sample R0",
  "Collapsed-R: raw eta0",
  sprintf("Collapsed-R: %.3f x eta0", collapsed_eta0_multiplier),
  "AIW-N: raw eta0",
  sprintf("AIW-N: %.3f x eta0", aiwn_eta0_multiplier)
))

ld_for_fit <- list(
  R_in, R0,
  collapsed_raw$selected_Rbar, collapsed_calibrated$selected_Rbar,
  aiwn_raw$selected_Rbar, aiwn_calibrated$selected_Rbar
)
plot_items <- Map(function(fit, R) list(object = fit, R = R),
                  fits, ld_for_fit)
names(plot_items) <- names(fits)

plot_iw_comparison(
  plot_items, causal = causal, ncol = 3L,
  main = "SuSiE-IW simulation: PIPs and reported credible sets"
)

# Compact numerical report printed after the figure.
summary_table <- do.call(rbind, Map(function(label, fit) {
  row <- summarize_iw_fit(fit, causal = causal)
  row$method <- label
  row
}, names(fits), fits))
rownames(summary_table) <- NULL
print(summary_table, row.names = FALSE, digits = 4)

cat("\nSelected mismatch parameters\n")
print(data.frame(
  method = c("Collapsed-R raw", "Collapsed-R calibrated",
             "AIW-N raw", "AIW-N calibrated"),
  lambda = c(collapsed_raw$selected_lambda,
             collapsed_calibrated$selected_lambda,
             aiwn_raw$selected_lambda, aiwn_calibrated$selected_lambda),
  raw_eta0 = c(collapsed_raw$raw_eta0,
               collapsed_calibrated$raw_eta0,
               aiwn_raw$selected_raw_eta0,
               aiwn_calibrated$selected_raw_eta0),
  eta0_used = c(collapsed_raw$eta0_used,
                collapsed_calibrated$eta0_used,
                aiwn_raw$selected_eta0,
                aiwn_calibrated$selected_eta0),
  row.names = NULL
), row.names = FALSE, digits = 5)
```
