---
title: "Coalescent SuSiE-IW simulation"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  run_analysis: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This page records the collaborator-facing coalescent SuSiE-IW simulation for
two and three causal effects. Reference LD is generated with `stdpopsim` and
`msprime`; in-sample LD and marginal associations are then generated under the
SuSiE-IW model.

Execution requires the coalescent Python dependencies and runs during ordinary
workflowr builds. Set `run_analysis: false` in the YAML header to skip the
complete simulation.

## Analysis

```{r analysis, eval=params$run_analysis}
release_dir <- if (file.exists("01-collapsed-R.R")) {
  "."
} else {
  "code"
}
source(file.path(release_dir, "01-collapsed-R.R"))
source(file.path(release_dir, "02-AIW-and-AIW-N.R"))
source(file.path(release_dir, "03-SER-fallback-and-plot.R"))

# -------------------------------------------------------------------------
# Settings used in the manuscript's coalescent SuSiE-IW experiment
# -------------------------------------------------------------------------

seed <- 7L
J <- 1000L
N0 <- 2000L                       # reference-panel sample size
coalescent_sample_size <- 2000L   # second population used for common-SNP QC
N <- 200000L                      # GWAS sample size
true_eta0 <- 200
h2 <- 1e-3
true_lambda <- 1e-3
true_L_values <- c(2L, 3L)
maximum_fitted_L <- 5L

lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)
collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

# -------------------------------------------------------------------------
# 1. Generate empirical reference LD R0 from a human coalescent model
# -------------------------------------------------------------------------

simulate_coalescent_R0 <- function(
    J, N0, second_population_size = N0, seed = 1L,
    chromosome = "chr22", start = 20e6, end = 21e6,
    split_time = 10L) {
  if (!requireNamespace("reticulate", quietly = TRUE)) {
    stop("Install the R package reticulate")
  }
  if (!nzchar(Sys.getenv("RETICULATE_PYTHON"))) {
    python <- Sys.which("python3")
    if (nzchar(python)) {
      Sys.setenv(RETICULATE_PYTHON = python)
      reticulate::use_python(python, required = FALSE)
    }
  }
  stdpopsim <- reticulate::import("stdpopsim")
  msprime <- reticulate::import("msprime")

  species <- stdpopsim$get_species("HomSap")
  contig <- species$get_contig(
    chromosome, left = as.integer(start), right = as.integer(end)
  )
  demography <- msprime$Demography()
  demography$add_population(name = "ANC", initial_size = 20000L)
  demography$add_population(name = "POP1", initial_size = 100000L)
  demography$add_population(name = "POP2", initial_size = 100000L)
  demography$add_population_split(
    time = as.integer(split_time),
    derived = list("POP1", "POP2"), ancestral = "ANC"
  )

  models <- list(
    msprime$DiscreteTimeWrightFisher(duration = 100L),
    msprime$StandardCoalescent()
  )
  tree_sequence <- msprime$sim_ancestry(
    samples = reticulate::dict(
      POP1 = as.integer(second_population_size),
      POP2 = as.integer(N0)
    ),
    demography = demography,
    recombination_rate = contig$recombination_map,
    model = models,
    random_seed = as.integer(seed)
  )
  tree_sequence <- msprime$sim_mutations(
    tree_sequence, rate = 1.29e-8, random_seed = 1L
  )

  population_ids <- list()
  for (population in reticulate::iterate(tree_sequence$populations())) {
    population_ids[[population$metadata$name]] <- population$id
  }
  genotype_haploid <- tree_sequence$genotype_matrix()
  to_diploid <- function(sample_ids) {
    columns <- as.integer(sample_ids) + 1L
    haplotypes <- genotype_haploid[, columns, drop = FALSE]
    genotypes <- haplotypes[, seq(1L, ncol(haplotypes), by = 2L)] +
      haplotypes[, seq(2L, ncol(haplotypes), by = 2L)]
    t(genotypes)
  }
  genotype_1 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP1)
  ))
  genotype_0 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP2)
  ))

  frequency_1 <- colMeans(genotype_1) / 2
  frequency_0 <- colMeans(genotype_0) / 2
  common <- frequency_1 > 0.05 & frequency_1 < 0.95 &
    frequency_0 > 0.05 & frequency_0 < 0.95
  genotype_1 <- genotype_1[, common, drop = FALSE]
  genotype_0 <- genotype_0[, common, drop = FALSE]
  if (ncol(genotype_0) < J) {
    stop("The coalescent region contains fewer than J common variants")
  }

  # Use the same seeded contiguous-block selection as the manuscript study.
  set.seed(seed)
  first_column <- if (ncol(genotype_0) == J) {
    1L
  } else {
    sample.int(ncol(genotype_0) - J + 1L, 1L)
  }
  columns <- first_column + seq_len(J) - 1L
  R0 <- stats::cor(genotype_0[, columns, drop = FALSE])
  R0 <- (R0 + t(R0)) / 2
  diag(R0) <- 1
  R0
}

R0 <- simulate_coalescent_R0(
  J = J, N0 = N0, second_population_size = coalescent_sample_size,
  seed = seed
)

# Plot the reference LD with a fixed color scale so heatmaps from different
# panel sizes or seeds are comparable.
old_par <- graphics::par(mar = c(4.2, 4.2, 3.2, 1.2))
ld_colors <- grDevices::colorRampPalette(
  c("#2166AC", "#F7F7F7", "#B2182B")
)(201L)
graphics::image(
  x = seq_len(J), y = seq_len(J), z = R0,
  zlim = c(-1, 1), col = ld_colors, useRaster = TRUE,
  asp = 1, xlab = "Variant index", ylab = "Variant index",
  main = sprintf("Coalescent reference LD R0 (N0 = %s)",
                 format(N0, big.mark = ","))
)
graphics::box()
graphics::par(old_par)

# -------------------------------------------------------------------------
# 2. SuSiE-IW draw: R | R0
#
#    Rbar_true = (1 - true_lambda) R0 + true_lambda I,
#    R_raw ~ IW_J(true_eta0 Rbar_true, true_eta0 + J + 1),
#    R = cov2cor(R_raw).
# -------------------------------------------------------------------------

Rbar_true <- (1 - true_lambda) * R0 + true_lambda * diag(J)
set.seed(2000000L + 10000L * seed + J + as.integer(true_eta0))
precision <- stats::rWishart(
  1L,
  df = true_eta0 + J + 1L,
  Sigma = solve(true_eta0 * Rbar_true)
)[, , 1L]
R <- stats::cov2cor(solve(precision))
R <- (R + t(R)) / 2
diag(R) <- 1

# Pick three separated, weak-LD variants. The L=2 architecture uses the first
# two; the L=3 architecture adds the third, so the architectures are nested.
choose_nested_weak_ld_variants <- function(R) {
  candidates <- unique(as.integer(round(seq(
    max(2, 0.08 * nrow(R)), min(nrow(R) - 1, 0.92 * nrow(R)),
    length.out = min(250L, nrow(R) - 2L)
  ))))
  minimum_separation <- max(8L, floor(nrow(R) / 12))
  first <- candidates[which.min(abs(candidates - round(nrow(R) / 4)))]
  available <- candidates[abs(candidates - first) >= minimum_separation]
  second <- available[which.min(abs(R[first, available]))]
  available <- candidates[
    abs(candidates - first) >= minimum_separation &
      abs(candidates - second) >= minimum_separation
  ]
  third <- available[which.min(vapply(available, function(j) {
    max(abs(R[j, c(first, second)]))
  }, numeric(1L)))]
  c(first, second, third)
}
causal_candidates <- choose_nested_weak_ld_variants(R)

# Use one shared noise draw so that the L=2 and L=3 examples differ only in b.
set.seed(2500000L + 10000L * seed + J + as.integer(true_eta0))
noise <- as.numeric(t(chol(R)) %*% stats::rnorm(J)) / sqrt(N)

# -------------------------------------------------------------------------
# 3. Generate beta and v | R, beta for L_true = 2 or 3
#
#    beta is proportional to (7,5) or (7,5,3), then scaled so beta' R beta=h2.
#    v | R,beta ~ N_J(R beta, R/N).
# -------------------------------------------------------------------------

simulate_architecture <- function(true_L) {
  stopifnot(true_L %in% c(2L, 3L))
  causal <- causal_candidates[seq_len(true_L)]
  weights <- c(7, 5, 3)[seq_len(true_L)]
  scale <- sqrt(
    h2 / drop(crossprod(
      weights, R[causal, causal, drop = FALSE] %*% weights
    ))
  )
  beta <- numeric(J)
  beta[causal] <- scale * weights
  v <- as.numeric(R %*% beta + noise)
  list(
    true_L = true_L,
    causal = causal,
    beta = beta,
    v = v,
    causal_LD = R[causal, causal, drop = FALSE],
    expected_z = sqrt(N) * as.numeric(
      R[causal, , drop = FALSE] %*% beta
    )
  )
}

simulation_L2 <- simulate_architecture(2L)
simulation_L3 <- simulate_architecture(3L)
simulations <- list(L2 = simulation_L2, L3 = simulation_L3)

# -------------------------------------------------------------------------
# 4. Fit the four manuscript methods
# -------------------------------------------------------------------------

fit_architecture <- function(simulation) {
  v <- simulation$v
  list(
    susie_in_sample = fit_susie_rss(
      v = v, R = R, N = N, L = maximum_fitted_L
    ),
    susie_reference = fit_susie_rss(
      v = v, R = R0, N = N, L = maximum_fitted_L
    ),
    collapsed_R = susie_iw(
      v = v, R0 = R0, N = N, L = maximum_fitted_L,
      lambda_grid = lambda_grid,
      eta0_multiplier = collapsed_eta0_multiplier,
      scaled_prior_variance = 0.2,
      estimate_prior_variance = TRUE,
      coverage = 0.95, min_abs_corr = 0.5,
      ser_fallback = TRUE,
      max_iter = 200L, tol = 1e-4, verbose = FALSE
    ),
    AIW_N = suppressMessages(susie_aiw(
      v = v, R0 = R0, N = N, L = maximum_fitted_L,
      approximation = "N",
      lambda_grid = lambda_grid,
      eta0_multiplier = aiwn_eta0_multiplier,
      scaled_prior_variance = 0.2,
      estimate_prior_variance = TRUE,
      coverage = 0.95, min_abs_corr = 0.5,
      ser_fallback = TRUE,
      max_iter = 500L, tol = 1e-3, verbose = FALSE
    ))
  )
}

fits_L2 <- fit_architecture(simulation_L2)
fits_L3 <- fit_architecture(simulation_L3)
fits <- list(L2 = fits_L2, L3 = fits_L3)

summarize_architecture <- function(label, simulation, fitted_methods) {
  table <- do.call(rbind, Map(function(method, fit) {
    row <- summarize_iw_fit(fit, causal = simulation$causal)
    row$architecture <- label
    row$method <- method
    row
  }, names(fitted_methods), fitted_methods))
  rownames(table) <- NULL
  table[, c("architecture", "method",
            setdiff(names(table), c("architecture", "method")))]
}

summary_table <- rbind(
  summarize_architecture("L_true = 2", simulation_L2, fits_L2),
  summarize_architecture("L_true = 3", simulation_L3, fits_L3)
)
print(summary_table, row.names = FALSE, digits = 4)

cat("\nGenerating parameters and causal variants\n")
print(data.frame(
  true_L = true_L_values,
  causal = vapply(simulations, function(x) {
    paste(x$causal, collapse = ",")
  }, character(1L)),
  expected_abs_z = vapply(simulations, function(x) {
    paste(round(abs(x$expected_z), 2), collapse = ",")
  }, character(1L)),
  max_abs_causal_LD = vapply(simulations, function(x) {
    max(abs(x$causal_LD[upper.tri(x$causal_LD)]))
  }, numeric(1L)),
  row.names = NULL
), row.names = FALSE)

cat("\nSelected mismatch parameters\n")
print(do.call(rbind, lapply(names(fits), function(label) {
  data.frame(
    architecture = label,
    method = c("collapsed-R", "AIW-N"),
    lambda = c(fits[[label]]$collapsed_R$selected_lambda,
               fits[[label]]$AIW_N$selected_lambda),
    raw_eta0 = c(fits[[label]]$collapsed_R$raw_eta0,
                 fits[[label]]$AIW_N$selected_raw_eta0),
    eta0_used = c(fits[[label]]$collapsed_R$eta0,
                  fits[[label]]$AIW_N$eta0),
    row.names = NULL
  )
})), row.names = FALSE, digits = 5)

# Render both the L=2 and L=3 comparison displays in the document.
for (label in names(fits)) {
  current_fits <- fits[[label]]
  plot_items <- list(
    "SuSiE-RSS: in-sample R" = list(
      object = current_fits$susie_in_sample, R = R
    ),
    "SuSiE-RSS: reference R0" = list(
      object = current_fits$susie_reference, R = R0
    ),
    "Collapsed-R" = list(
      object = current_fits$collapsed_R,
      R = current_fits$collapsed_R$selected_Rbar
    ),
    "AIW-N" = list(
      object = current_fits$AIW_N,
      R = current_fits$AIW_N$selected_Rbar
    )
  )
  plot_iw_comparison(
    plot_items, causal = simulations[[label]]$causal, ncol = 2L,
    main = paste0("Coalescent SuSiE-IW simulation: ", label)
  )
}
```
