---
title: "Comparison with a null posterior-mean LD estimate"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  run_analysis: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This page records the comparison proposed by Peter Carbonetto. The null model
first estimates the inverse-Wishart concentration without sparse effects,
constructs its posterior-mean LD matrix, and then fits SuSiE-RSS. The simulation
compares its concentration estimates, coverage, and power with collapsed-R and
AIW-N.

The default experiment contains multiple coalescent simulations and runs during
ordinary workflowr builds. Set `run_analysis: false` in the YAML header to skip
it.

## Analysis

```{r analysis, eval=params$run_analysis}
# Compare Peter Carbonetto's null-posterior LD estimate with collapsed-R and
# AIW-N in the coalescent SuSiE-IW simulation. Run this file from either the
# repository root or code/. It writes two figures, a CSV and
# an RDS file beside this script, and also draws the figures in RStudio.

release_dir <- if (file.exists("01-collapsed-R.R")) {
  "."
} else {
  "code"
}
source(file.path(release_dir, "01-collapsed-R.R"))
source(file.path(release_dir, "02-AIW-and-AIW-N.R"))
source(file.path(release_dir, "03-SER-fallback-and-plot.R"))

# -------------------------------------------------------------------------
# Settings
# -------------------------------------------------------------------------

seed <- 7L
J <- as.integer(Sys.getenv("NULL_R_J", "500"))
number_of_replications <- as.integer(Sys.getenv("NULL_R_REPS", "20"))
N0 <- 2000L
coalescent_sample_size <- 2000L
N <- 200000L
h2 <- 1e-3
true_lambda <- 1e-3
true_eta0_grid <- c(25, 50, 100, 200, 400, 800)
true_L <- 2L
effect_weights <- c(7, 5)
maximum_fitted_L <- 5L
lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)

# Frozen calibration used by the current operational procedures. Their raw
# eta0 estimates are retained separately and are what the eta0 figure shows.
collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

eta0_bounds <- c(1e-4, 1e8)
eta0_grid_size <- 121L
bootstrap_replicates <- 1000L
output_stem <- file.path("experiments", "03-null-estimate-R")

stopifnot(
  J >= 20L, number_of_replications >= 1L, true_L == 2L,
  length(effect_weights) == true_L
)

# -------------------------------------------------------------------------
# 1. One empirical reference-LD matrix from a human coalescent model
# -------------------------------------------------------------------------

simulate_coalescent_R0 <- function(
    J, N0, second_population_size = N0, seed = 1L,
    chromosome = "chr22", start = 20e6, end = 21e6,
    split_time = 10L) {
  if (!requireNamespace("reticulate", quietly = TRUE)) {
    stop("Install the R package reticulate")
  }
  if (!nzchar(Sys.getenv("RETICULATE_PYTHON"))) {
    python <- Sys.which("python3")
    if (nzchar(python)) {
      Sys.setenv(RETICULATE_PYTHON = python)
      reticulate::use_python(python, required = FALSE)
    }
  }
  stdpopsim <- reticulate::import("stdpopsim")
  msprime <- reticulate::import("msprime")

  species <- stdpopsim$get_species("HomSap")
  contig <- species$get_contig(
    chromosome, left = as.integer(start), right = as.integer(end)
  )
  demography <- msprime$Demography()
  demography$add_population(name = "ANC", initial_size = 20000L)
  demography$add_population(name = "POP1", initial_size = 100000L)
  demography$add_population(name = "POP2", initial_size = 100000L)
  demography$add_population_split(
    time = as.integer(split_time),
    derived = list("POP1", "POP2"), ancestral = "ANC"
  )
  models <- list(
    msprime$DiscreteTimeWrightFisher(duration = 100L),
    msprime$StandardCoalescent()
  )
  tree_sequence <- msprime$sim_ancestry(
    samples = reticulate::dict(
      POP1 = as.integer(second_population_size),
      POP2 = as.integer(N0)
    ),
    demography = demography,
    recombination_rate = contig$recombination_map,
    model = models,
    random_seed = as.integer(seed)
  )
  tree_sequence <- msprime$sim_mutations(
    tree_sequence, rate = 1.29e-8, random_seed = 1L
  )

  population_ids <- list()
  for (population in reticulate::iterate(tree_sequence$populations())) {
    population_ids[[population$metadata$name]] <- population$id
  }
  genotype_haploid <- tree_sequence$genotype_matrix()
  to_diploid <- function(sample_ids) {
    columns <- as.integer(sample_ids) + 1L
    haplotypes <- genotype_haploid[, columns, drop = FALSE]
    genotypes <- haplotypes[, seq(1L, ncol(haplotypes), by = 2L)] +
      haplotypes[, seq(2L, ncol(haplotypes), by = 2L)]
    t(genotypes)
  }
  genotype_1 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP1)
  ))
  genotype_0 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP2)
  ))
  frequency_1 <- colMeans(genotype_1) / 2
  frequency_0 <- colMeans(genotype_0) / 2
  common <- frequency_1 > 0.05 & frequency_1 < 0.95 &
    frequency_0 > 0.05 & frequency_0 < 0.95
  genotype_0 <- genotype_0[, common, drop = FALSE]
  if (ncol(genotype_0) < J) {
    stop("The coalescent region contains fewer than J common variants")
  }
  set.seed(seed)
  first_column <- if (ncol(genotype_0) == J) {
    1L
  } else {
    sample.int(ncol(genotype_0) - J + 1L, 1L)
  }
  columns <- first_column + seq_len(J) - 1L
  R0 <- stats::cor(genotype_0[, columns, drop = FALSE])
  R0 <- (R0 + t(R0)) / 2
  diag(R0) <- 1
  R0
}

# -------------------------------------------------------------------------
# 2. Null-model eta0 profile and posterior-mean R
#
# R ~ IW_J(eta0 Rbar, eta0 + J + 1),  v | R ~ N_J(0, R/N).
# Therefore
# v ~ t_{eta0+2}(0, eta0 Rbar / {N(eta0+2)}),
# and E(R | v) = (eta0 Rbar + N v v') / (eta0 + 1).
# -------------------------------------------------------------------------

null_log_marginal_eta0 <- function(eta0, v, Rbar_chol, N) {
  if (length(eta0) != 1L || !is.finite(eta0) || eta0 <= 0) {
    return(-Inf)
  }
  J <- length(v)
  transformed_v <- backsolve(Rbar_chol, v, transpose = TRUE)
  quadratic <- N * sum(transformed_v^2)
  log_determinant <- 2 * sum(log(diag(Rbar_chol)))

  # For even J this finite sum is more stable than subtracting two lgamma
  # values when eta0 is large. The fallback handles arbitrary J.
  half_J <- J / 2
  if (half_J == floor(half_J)) {
    log_gamma_ratio <- sum(log((eta0 + 2) / 2 + seq.int(0, half_J - 1)))
  } else {
    log_gamma_ratio <- lgamma((eta0 + J + 2) / 2) -
      lgamma((eta0 + 2) / 2)
  }
  log_gamma_ratio + J / 2 * log(N / (pi * eta0)) -
    log_determinant / 2 -
    (eta0 + J + 2) / 2 * log1p(quadratic / eta0)
}

estimate_null_eta0 <- function(
    v, Rbar, N, bounds = c(1e-4, 1e8), grid_size = 121L) {
  Rbar_chol <- chol((Rbar + t(Rbar)) / 2)
  log_grid <- seq(log(bounds[1L]), log(bounds[2L]), length.out = grid_size)
  objective <- function(log_eta0) {
    null_log_marginal_eta0(exp(log_eta0), v, Rbar_chol, N)
  }
  grid_objective <- vapply(log_grid, objective, numeric(1L))
  best <- which.max(grid_objective)
  lower <- log_grid[max(1L, best - 1L)]
  upper <- log_grid[min(length(log_grid), best + 1L)]
  refined <- if (lower < upper) {
    optimize(objective, interval = c(lower, upper), maximum = TRUE)
  } else {
    list(maximum = log_grid[best], objective = grid_objective[best])
  }
  candidates <- data.frame(
    log_eta0 = c(log_grid[1L], refined$maximum, tail(log_grid, 1L)),
    objective = c(grid_objective[1L], refined$objective,
                  tail(grid_objective, 1L))
  )
  selected <- which.max(candidates$objective)
  list(
    eta0 = exp(candidates$log_eta0[selected]),
    log_marginal = candidates$objective[selected],
    boundary = selected != 2L,
    Rbar_chol = Rbar_chol
  )
}

fit_null_estimate_R <- function(
    v, R0, N, L = 5L, lambda_grid = NULL,
    eta0_bounds = NULL, eta0_grid_size = NULL,
    prior_variance = 0.2, max_iter = 500L, tol = 1e-3,
    verbose = FALSE) {
  if (is.null(lambda_grid)) {
    lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)
  }
  if (is.null(eta0_bounds)) eta0_bounds <- c(1e-4, 1e8)
  if (is.null(eta0_grid_size)) eta0_grid_size <- 121L
  J <- length(v)
  candidate_fits <- vector("list", length(lambda_grid))
  profile <- vector("list", length(lambda_grid))
  for (i in seq_along(lambda_grid)) {
    lambda <- lambda_grid[i]
    Rbar <- (1 - lambda) * R0 + lambda * diag(J)
    eta_fit <- estimate_null_eta0(
      v, Rbar, N, bounds = eta0_bounds, grid_size = eta0_grid_size
    )
    R_hat <- (eta_fit$eta0 * Rbar + N * tcrossprod(v)) /
      (eta_fit$eta0 + 1)
    R_hat <- (R_hat + t(R_hat)) / 2

    # This is Peter's posterior mean exactly as proposed. In particular, it
    # is not converted with cov2cor before the SuSiE-RSS refit.
    fit <- fit_susie_rss(
      v = v, R = R_hat, N = N, L = L,
      prior_variance = prior_variance, max_iter = max_iter, tol = tol
    )
    susie_elbo <- tail(fit$elbo, 1L)
    candidate_fits[[i]] <- fit
    profile[[i]] <- data.frame(
      lambda = lambda,
      eta0 = eta_fit$eta0,
      null_log_marginal = eta_fit$log_marginal,
      eta0_boundary = eta_fit$boundary,
      susie_elbo = susie_elbo,
      converged = isTRUE(fit$converged)
    )
    if (verbose) {
      message(sprintf(
        "null-R: lambda=%g eta0=%.5g SuSiE ELBO=%.6f",
        lambda, eta_fit$eta0, susie_elbo
      ))
    }
  }
  profile <- do.call(rbind, profile)
  best <- which.max(profile$susie_elbo)
  result <- candidate_fits[[best]]
  result$method <- "Null posterior-mean R"
  result$selected_lambda <- profile$lambda[best]
  result$raw_eta0 <- profile$eta0[best]
  result$eta0 <- profile$eta0[best]
  result$eta0_used <- profile$eta0[best]
  result$R_hat <- (
    result$eta0 * ((1 - result$selected_lambda) * R0 +
                     result$selected_lambda * diag(J)) + N * tcrossprod(v)
  ) / (result$eta0 + 1)
  result$profile <- profile
  class(result) <- unique(c("null_estimate_R", class(result)))
  result
}

# -------------------------------------------------------------------------
# 3. SuSiE-IW simulation with two weak-LD causal effects in a 7:5 ratio
# -------------------------------------------------------------------------

choose_weak_ld_variants <- function(R) {
  candidates <- unique(as.integer(round(seq(
    max(2, 0.08 * nrow(R)), min(nrow(R) - 1, 0.92 * nrow(R)),
    length.out = min(250L, nrow(R) - 2L)
  ))))
  minimum_separation <- max(8L, floor(nrow(R) / 12))
  first <- candidates[which.min(abs(candidates - round(nrow(R) / 4)))]
  available <- candidates[abs(candidates - first) >= minimum_separation]
  second <- available[which.min(abs(R[first, available]))]
  c(first, second)
}

simulate_dataset <- function(true_eta0, replication) {
  simulation_seed <- seed * 10000000L + as.integer(true_eta0) * 100L +
    replication
  set.seed(simulation_seed)
  precision <- stats::rWishart(
    1L, df = true_eta0 + J + 1L,
    Sigma = Rbar_true_inverse / true_eta0
  )[, , 1L]
  R <- stats::cov2cor(chol2inv(chol(precision)))
  R <- (R + t(R)) / 2
  diag(R) <- 1

  causal <- choose_weak_ld_variants(R)
  beta_scale <- sqrt(
    h2 / drop(crossprod(
      effect_weights, R[causal, causal, drop = FALSE] %*% effect_weights
    ))
  )
  beta <- numeric(J)
  beta[causal] <- beta_scale * effect_weights
  noise <- as.numeric(t(chol(R)) %*% stats::rnorm(J)) / sqrt(N)
  v <- as.numeric(R %*% beta + noise)
  list(
    v = v, R = R, beta = beta, causal = causal,
    causal_LD = R[causal[1L], causal[2L]],
    expected_z = sqrt(N) * as.numeric(R[causal, , drop = FALSE] %*% beta)
  )
}

evaluate_fit <- function(fit, causal) {
  sets <- susie_get_reported_cs(fit)
  covered <- vapply(causal, function(j) {
    any(vapply(sets, function(set) j %in% set, logical(1L)))
  }, logical(1L))
  true_sets <- sum(vapply(sets, function(set) any(set %in% causal), logical(1L)))
  data.frame(
    estimated_eta0 = if (!is.null(fit$raw_eta0)) {
      fit$raw_eta0
    } else {
      fit$selected_raw_eta0
    },
    eta0_used = if (!is.null(fit$eta0_used)) fit$eta0_used else fit$eta0,
    selected_lambda = fit$selected_lambda,
    reported_L = length(sets),
    true_sets = true_sets,
    false_sets = length(sets) - true_sets,
    causal_power = mean(covered),
    all_causal_covered = all(covered),
    fallback = isTRUE(fit$postprocess$fallback_used),
    converged = isTRUE(fit$converged)
  )
}

if (!identical(Sys.getenv("NULL_R_FUNCTIONS_ONLY"), "1")) {
method_levels <- c("Null posterior-mean R", "collapsed-R", "AIW-N")
reuse_results <- identical(Sys.getenv("NULL_R_REUSE_RESULTS"), "1")
if (reuse_results) {
  saved_results <- readRDS(paste0(output_stem, ".rds"))
  if (!identical(as.numeric(saved_results$settings$true_eta0_grid),
                 as.numeric(true_eta0_grid)) ||
      saved_results$settings$J != J ||
      saved_results$settings$replications != number_of_replications) {
    stop("Saved results do not match the current eta0 grid, J and replications")
  }
  message("Reusing completed results and regenerating summaries and figures")
  R0 <- saved_results$R0
  results <- saved_results$results
} else {
message("Generating the coalescent reference LD matrix")
R0 <- simulate_coalescent_R0(
  J = J, N0 = N0, second_population_size = coalescent_sample_size,
  seed = seed
)
Rbar_true <- (1 - true_lambda) * R0 + true_lambda * diag(J)
Rbar_true_inverse <- chol2inv(chol(Rbar_true))

result_rows <- vector(
  "list", length(true_eta0_grid) * number_of_replications * 3L
)
row_index <- 0L
for (true_eta0 in true_eta0_grid) {
  for (replication in seq_len(number_of_replications)) {
    message(sprintf(
      "true eta0=%s, replication %d/%d",
      format(true_eta0, scientific = FALSE), replication,
      number_of_replications
    ))
    simulation <- simulate_dataset(true_eta0, replication)
    fits <- list(
      "Null posterior-mean R" = fit_null_estimate_R(
        v = simulation$v, R0 = R0, N = N, L = maximum_fitted_L,
        lambda_grid = lambda_grid, verbose = FALSE
      ),
      "collapsed-R" = susie_iw(
        v = simulation$v, R0 = R0, N = N, L = maximum_fitted_L,
        lambda_grid = lambda_grid,
        eta0_multiplier = collapsed_eta0_multiplier,
        scaled_prior_variance = 0.2, estimate_prior_variance = TRUE,
        coverage = 0.95, min_abs_corr = 0.5,
        ser_fallback = TRUE, partition_ser = FALSE,
        max_iter = 200L, tol = 1e-4, verbose = FALSE
      ),
      "AIW-N" = suppressMessages(susie_aiw(
        v = simulation$v, R0 = R0, N = N, L = maximum_fitted_L,
        approximation = "N", lambda_grid = lambda_grid,
        eta0_multiplier = aiwn_eta0_multiplier,
        scaled_prior_variance = 0.2, estimate_prior_variance = TRUE,
        coverage = 0.95, min_abs_corr = 0.5,
        ser_fallback = TRUE, partition_ser = FALSE,
        max_iter = 500L, tol = 1e-3, verbose = FALSE
      ))
    )
    for (method in names(fits)) {
      row_index <- row_index + 1L
      row <- evaluate_fit(fits[[method]], simulation$causal)
      row$true_eta0 <- true_eta0
      row$replication <- replication
      row$method <- method
      row$causal_LD <- simulation$causal_LD
      row$expected_z_stronger <- max(abs(simulation$expected_z))
      row$expected_z_weaker <- min(abs(simulation$expected_z))
      result_rows[[row_index]] <- row
    }
  }
}
results <- do.call(rbind, result_rows)
}
results$method <- factor(results$method, levels = method_levels)

# -------------------------------------------------------------------------
# 4. Summaries and figures
# -------------------------------------------------------------------------

summarize_cell <- function(data, bootstrap_replicates = 1000L) {
  metrics <- function(x) c(
    coverage = if (sum(x$reported_L) > 0) {
      sum(x$true_sets) / sum(x$reported_L)
    } else {
      NA_real_
    },
    power = mean(x$causal_power)
  )
  estimate <- metrics(data)
  bootstrap <- replicate(bootstrap_replicates, {
    metrics(data[sample.int(nrow(data), nrow(data), replace = TRUE), ])
  })
  interval <- function(values, probability) {
    values <- values[is.finite(values)]
    if (length(values)) {
      unname(quantile(values, probability, names = FALSE))
    } else {
      NA_real_
    }
  }
  data.frame(
    coverage = estimate["coverage"],
    coverage_lower = interval(bootstrap["coverage", ], 0.025),
    coverage_upper = interval(bootstrap["coverage", ], 0.975),
    power = estimate["power"],
    power_lower = interval(bootstrap["power", ], 0.025),
    power_upper = interval(bootstrap["power", ], 0.975)
  )
}

summary_rows <- list()
summary_index <- 0L
for (eta0 in true_eta0_grid) {
  for (method in method_levels) {
    summary_index <- summary_index + 1L
    cell <- results[
      results$true_eta0 == eta0 & as.character(results$method) == method,
    ]
    set.seed(900000L + as.integer(eta0) + match(method, method_levels))
    row <- summarize_cell(cell, bootstrap_replicates)
    row$true_eta0 <- eta0
    row$method <- method
    row$median_estimated_eta0 <- median(cell$estimated_eta0)
    row$eta0_q25 <- unname(quantile(cell$estimated_eta0, 0.25))
    row$eta0_q75 <- unname(quantile(cell$estimated_eta0, 0.75))
    row$mean_reported_L <- mean(cell$reported_L)
    row$mean_false_sets <- mean(cell$false_sets)
    row$fallback_frequency <- mean(cell$fallback)
    summary_rows[[summary_index]] <- row
  }
}
summary_table <- do.call(rbind, summary_rows)
summary_table$method <- factor(summary_table$method, levels = method_levels)

method_colors <- c("#D73027", "#2CA02C", "#6A3D9A")
method_shapes <- c(17, 16, 18)
names(method_colors) <- names(method_shapes) <- method_levels

draw_eta0 <- function() {
  normal_limit <- is.infinite(results$estimated_eta0) |
    results$estimated_eta0 >= 0.99 * eta0_bounds[2L]
  finite_eta0 <- results$estimated_eta0[
    is.finite(results$estimated_eta0) & !normal_limit
  ]
  display_ceiling <- max(
    2 * max(true_eta0_grid), 1.25 * max(finite_eta0)
  )
  display_eta0 <- results$estimated_eta0
  display_eta0[normal_limit] <- display_ceiling
  limits <- range(c(true_eta0_grid, finite_eta0, display_eta0))
  plot(
    true_eta0_grid, true_eta0_grid, type = "n", log = "xy",
    xlim = range(true_eta0_grid), ylim = limits,
    xlab = expression("true " * eta[0]),
    ylab = expression("raw estimated " * hat(eta)[0]),
    main = expression("Recovery of " * eta[0]), xaxt = "n"
  )
  axis(1, at = true_eta0_grid,
       labels = format(true_eta0_grid, big.mark = ",", scientific = FALSE))
  grid(col = "grey88", lty = 3)
  abline(0, 1, lty = 2, col = "grey35")
  for (method in method_levels) {
    individual <- results[as.character(results$method) == method, ]
    individual_eta0 <- pmin(individual$estimated_eta0, display_ceiling)
    points(
      individual$true_eta0, individual_eta0,
      pch = method_shapes[method], col = grDevices::adjustcolor(
        method_colors[method], alpha.f = 0.22
      ), cex = 0.72
    )
    medians <- summary_table[as.character(summary_table$method) == method, ]
    median_eta0 <- pmin(medians$median_estimated_eta0, display_ceiling)
    eta0_q25 <- pmin(medians$eta0_q25, display_ceiling)
    eta0_q75 <- pmin(medians$eta0_q75, display_ceiling)
    lines(
      medians$true_eta0, median_eta0,
      type = "b", pch = method_shapes[method], lwd = 2.2,
      col = method_colors[method]
    )
    show_interval <- is.finite(eta0_q25) & is.finite(eta0_q75) &
      eta0_q25 < eta0_q75
    if (any(show_interval)) {
      arrows(
        medians$true_eta0[show_interval], eta0_q25[show_interval],
        medians$true_eta0[show_interval], eta0_q75[show_interval],
        angle = 90, code = 3, length = 0.03, col = method_colors[method]
      )
    }
  }
  legend(
    "topleft", legend = method_levels, col = method_colors,
    pch = method_shapes, lwd = 2, bty = "n", cex = 0.83
  )
  if (any(normal_limit)) {
    mtext("Normal-limit estimates are shown at the upper plotting boundary",
          side = 3, line = 0.15, cex = 0.7, adj = 1)
  }
}

draw_performance <- function() {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)
  par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3.2, 1), las = 1)
  for (metric in c("coverage", "power")) {
    plot(
      true_eta0_grid, rep(NA_real_, length(true_eta0_grid)),
      type = "n", log = "x", ylim = c(0, 1.03), xaxt = "n",
      xlab = expression("true " * eta[0]),
      ylab = if (metric == "coverage") {
        "credible-set coverage"
      } else {
        "causal-variant power"
      },
      main = if (metric == "coverage") "Coverage" else "Power"
    )
    axis(1, at = true_eta0_grid,
         labels = format(true_eta0_grid, big.mark = ",", scientific = FALSE))
    grid(col = "grey88", lty = 3)
    for (method in method_levels) {
      data <- summary_table[as.character(summary_table$method) == method, ]
      lines(
        data$true_eta0, data[[metric]], type = "b",
        pch = method_shapes[method], lwd = 2.2, col = method_colors[method]
      )
      lower <- data[[paste0(metric, "_lower")]]
      upper <- data[[paste0(metric, "_upper")]]
      show_interval <- is.finite(lower) & is.finite(upper) & lower < upper
      if (any(show_interval)) {
        arrows(
          data$true_eta0[show_interval], lower[show_interval],
          data$true_eta0[show_interval], upper[show_interval],
          angle = 90, code = 3, length = 0.03, col = method_colors[method]
        )
      }
    }
    if (metric == "coverage") {
      null_summary <- summary_table[
        as.character(summary_table$method) == "Null posterior-mean R", 
      ]
      no_sets <- !is.finite(null_summary$coverage)
      if (any(no_sets)) {
        points(
          null_summary$true_eta0[no_sets], rep(0.025, sum(no_sets)),
          pch = 4, lwd = 1.7, col = method_colors["Null posterior-mean R"]
        )
        text(
          null_summary$true_eta0[no_sets], rep(0.065, sum(no_sets)),
          labels = "no CS", cex = 0.68,
          col = method_colors["Null posterior-mean R"]
        )
      }
    }
    if (metric == "power") {
      legend(
        "bottomright", legend = method_levels, col = method_colors,
        pch = method_shapes, lwd = 2, bty = "n", cex = 0.78
      )
    }
  }
}

write_figure <- function(filename, width, height, draw) {
  grDevices::png(
    paste0(filename, ".png"), width = width * 180,
    height = height * 180, res = 180
  )
  draw()
  grDevices::dev.off()
  grDevices::pdf(paste0(filename, ".pdf"), width = width, height = height)
  draw()
  grDevices::dev.off()
}

write_figure(paste0(output_stem, "-estimated-eta0"), 6.5, 5.3, draw_eta0)
write_figure(paste0(output_stem, "-coverage-power"), 10, 4.7,
             draw_performance)
utils::write.csv(results, paste0(output_stem, "-results.csv"), row.names = FALSE)
utils::write.csv(
  summary_table, paste0(output_stem, "-summary.csv"), row.names = FALSE
)
saveRDS(
  list(
    settings = list(
      seed = seed, J = J, replications = number_of_replications,
      N0 = N0, N = N, h2 = h2, true_lambda = true_lambda,
      true_eta0_grid = true_eta0_grid, true_L = true_L,
      effect_weights = effect_weights, lambda_grid = lambda_grid,
      collapsed_eta0_multiplier = collapsed_eta0_multiplier,
      aiwn_eta0_multiplier = aiwn_eta0_multiplier
    ),
    R0 = R0, results = results, summary = summary_table
  ),
  paste0(output_stem, ".rds")
)

cat("\nSummary\n")
print(summary_table, row.names = FALSE, digits = 4)
cat("\nMedian absolute causal LD:", median(abs(results$causal_LD)), "\n")
cat("Expected |z| range:",
    range(c(results$expected_z_weaker, results$expected_z_stronger)), "\n")

draw_eta0()
draw_performance()
}
```
