---
title: "Single-dataset null-estimate-R demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  run_analysis: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This page records the single-dataset companion to the null posterior-mean LD
comparison. It exposes `J`, the true concentration, and the replication index
as simple settings, then displays fitted PIPs and estimated concentrations for
the reference methods.

The example is generated and fitted during ordinary workflowr builds. Set
`run_analysis: false` in the YAML header to skip it.

## Analysis

```{r analysis, eval=params$run_analysis}
# Compare Peter Carbonetto's null-posterior LD estimate with collapsed-R and
# AIW-N in the coalescent SuSiE-IW simulation. Run this file from either the
# repository root or code/. It writes two figures, a CSV and
# an RDS file beside this script, and also draws the figures in RStudio.

release_dir <- if (file.exists("01-collapsed-R.R")) {
  "."
} else {
  "code"
}
source(file.path(release_dir, "01-collapsed-R.R"))
source(file.path(release_dir, "02-AIW-and-AIW-N.R"))
source(file.path(release_dir, "03-SER-fallback-and-plot.R"))

# -------------------------------------------------------------------------
# Settings
# -------------------------------------------------------------------------

seed <- 7L
J <- as.integer(Sys.getenv("NULL_R_J", "500"))
number_of_replications <- as.integer(Sys.getenv("NULL_R_REPS", "20"))
N0 <- 2000L
coalescent_sample_size <- 2000L
N <- 200000L
h2 <- 1e-3
true_lambda <- 1e-3
true_eta0_grid <- c(25, 50, 100, 200, 400, 800)
true_L <- 2L
effect_weights <- c(7, 5)
maximum_fitted_L <- 5L
lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)

# Frozen calibration used by the current operational procedures. Their raw
# eta0 estimates are retained separately and are what the eta0 figure shows.
collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

eta0_bounds <- c(1e-4, 1e8)
eta0_grid_size <- 121L
bootstrap_replicates <- 1000L
output_stem <- file.path("experiments", "03-null-estimate-R")

stopifnot(
  J >= 20L, number_of_replications >= 1L, true_L == 2L,
  length(effect_weights) == true_L
)

# -------------------------------------------------------------------------
# 1. One empirical reference-LD matrix from a human coalescent model
# -------------------------------------------------------------------------

simulate_coalescent_R0 <- function(
    J, N0, second_population_size = N0, seed = 1L,
    chromosome = "chr22", start = 20e6, end = 21e6,
    split_time = 10L) {
  if (!requireNamespace("reticulate", quietly = TRUE)) {
    stop("Install the R package reticulate")
  }
  if (!nzchar(Sys.getenv("RETICULATE_PYTHON"))) {
    python <- Sys.which("python3")
    if (nzchar(python)) {
      Sys.setenv(RETICULATE_PYTHON = python)
      reticulate::use_python(python, required = FALSE)
    }
  }
  stdpopsim <- reticulate::import("stdpopsim")
  msprime <- reticulate::import("msprime")

  species <- stdpopsim$get_species("HomSap")
  contig <- species$get_contig(
    chromosome, left = as.integer(start), right = as.integer(end)
  )
  demography <- msprime$Demography()
  demography$add_population(name = "ANC", initial_size = 20000L)
  demography$add_population(name = "POP1", initial_size = 100000L)
  demography$add_population(name = "POP2", initial_size = 100000L)
  demography$add_population_split(
    time = as.integer(split_time),
    derived = list("POP1", "POP2"), ancestral = "ANC"
  )
  models <- list(
    msprime$DiscreteTimeWrightFisher(duration = 100L),
    msprime$StandardCoalescent()
  )
  tree_sequence <- msprime$sim_ancestry(
    samples = reticulate::dict(
      POP1 = as.integer(second_population_size),
      POP2 = as.integer(N0)
    ),
    demography = demography,
    recombination_rate = contig$recombination_map,
    model = models,
    random_seed = as.integer(seed)
  )
  tree_sequence <- msprime$sim_mutations(
    tree_sequence, rate = 1.29e-8, random_seed = 1L
  )

  population_ids <- list()
  for (population in reticulate::iterate(tree_sequence$populations())) {
    population_ids[[population$metadata$name]] <- population$id
  }
  genotype_haploid <- tree_sequence$genotype_matrix()
  to_diploid <- function(sample_ids) {
    columns <- as.integer(sample_ids) + 1L
    haplotypes <- genotype_haploid[, columns, drop = FALSE]
    genotypes <- haplotypes[, seq(1L, ncol(haplotypes), by = 2L)] +
      haplotypes[, seq(2L, ncol(haplotypes), by = 2L)]
    t(genotypes)
  }
  genotype_1 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP1)
  ))
  genotype_0 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP2)
  ))
  frequency_1 <- colMeans(genotype_1) / 2
  frequency_0 <- colMeans(genotype_0) / 2
  common <- frequency_1 > 0.05 & frequency_1 < 0.95 &
    frequency_0 > 0.05 & frequency_0 < 0.95
  genotype_0 <- genotype_0[, common, drop = FALSE]
  if (ncol(genotype_0) < J) {
    stop("The coalescent region contains fewer than J common variants")
  }
  set.seed(seed)
  first_column <- if (ncol(genotype_0) == J) {
    1L
  } else {
    sample.int(ncol(genotype_0) - J + 1L, 1L)
  }
  columns <- first_column + seq_len(J) - 1L
  R0 <- stats::cor(genotype_0[, columns, drop = FALSE])
  R0 <- (R0 + t(R0)) / 2
  diag(R0) <- 1
  R0
}

# -------------------------------------------------------------------------
# 2. Null-model eta0 profile and posterior-mean R
#
# R ~ IW_J(eta0 Rbar, eta0 + J + 1),  v | R ~ N_J(0, R/N).
# Therefore
# v ~ t_{eta0+2}(0, eta0 Rbar / {N(eta0+2)}),
# and E(R | v) = (eta0 Rbar + N v v') / (eta0 + 1).
# -------------------------------------------------------------------------

null_log_marginal_eta0 <- function(eta0, v, Rbar_chol, N) {
  if (length(eta0) != 1L || !is.finite(eta0) || eta0 <= 0) {
    return(-Inf)
  }
  J <- length(v)
  transformed_v <- backsolve(Rbar_chol, v, transpose = TRUE)
  quadratic <- N * sum(transformed_v^2)
  log_determinant <- 2 * sum(log(diag(Rbar_chol)))

  # For even J this finite sum is more stable than subtracting two lgamma
  # values when eta0 is large. The fallback handles arbitrary J.
  half_J <- J / 2
  if (half_J == floor(half_J)) {
    log_gamma_ratio <- sum(log((eta0 + 2) / 2 + seq.int(0, half_J - 1)))
  } else {
    log_gamma_ratio <- lgamma((eta0 + J + 2) / 2) -
      lgamma((eta0 + 2) / 2)
  }
  log_gamma_ratio + J / 2 * log(N / (pi * eta0)) -
    log_determinant / 2 -
    (eta0 + J + 2) / 2 * log1p(quadratic / eta0)
}

estimate_null_eta0 <- function(
    v, Rbar, N, bounds = c(1e-4, 1e8), grid_size = 121L) {
  Rbar_chol <- chol((Rbar + t(Rbar)) / 2)
  log_grid <- seq(log(bounds[1L]), log(bounds[2L]), length.out = grid_size)
  objective <- function(log_eta0) {
    null_log_marginal_eta0(exp(log_eta0), v, Rbar_chol, N)
  }
  grid_objective <- vapply(log_grid, objective, numeric(1L))
  best <- which.max(grid_objective)
  lower <- log_grid[max(1L, best - 1L)]
  upper <- log_grid[min(length(log_grid), best + 1L)]
  refined <- if (lower < upper) {
    optimize(objective, interval = c(lower, upper), maximum = TRUE)
  } else {
    list(maximum = log_grid[best], objective = grid_objective[best])
  }
  candidates <- data.frame(
    log_eta0 = c(log_grid[1L], refined$maximum, tail(log_grid, 1L)),
    objective = c(grid_objective[1L], refined$objective,
                  tail(grid_objective, 1L))
  )
  selected <- which.max(candidates$objective)
  list(
    eta0 = exp(candidates$log_eta0[selected]),
    log_marginal = candidates$objective[selected],
    boundary = selected != 2L,
    Rbar_chol = Rbar_chol
  )
}

fit_null_estimate_R <- function(
    v, R0, N, L = 5L, lambda_grid = NULL,
    eta0_bounds = NULL, eta0_grid_size = NULL,
    prior_variance = 0.2, max_iter = 500L, tol = 1e-3,
    verbose = FALSE) {
  if (is.null(lambda_grid)) {
    lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)
  }
  if (is.null(eta0_bounds)) eta0_bounds <- c(1e-4, 1e8)
  if (is.null(eta0_grid_size)) eta0_grid_size <- 121L
  J <- length(v)
  candidate_fits <- vector("list", length(lambda_grid))
  profile <- vector("list", length(lambda_grid))
  for (i in seq_along(lambda_grid)) {
    lambda <- lambda_grid[i]
    Rbar <- (1 - lambda) * R0 + lambda * diag(J)
    eta_fit <- estimate_null_eta0(
      v, Rbar, N, bounds = eta0_bounds, grid_size = eta0_grid_size
    )
    R_hat <- (eta_fit$eta0 * Rbar + N * tcrossprod(v)) /
      (eta_fit$eta0 + 1)
    R_hat <- (R_hat + t(R_hat)) / 2

    # This is Peter's posterior mean exactly as proposed. In particular, it
    # is not converted with cov2cor before the SuSiE-RSS refit.
    fit <- fit_susie_rss(
      v = v, R = R_hat, N = N, L = L,
      prior_variance = prior_variance, max_iter = max_iter, tol = tol
    )
    susie_elbo <- tail(fit$elbo, 1L)
    candidate_fits[[i]] <- fit
    profile[[i]] <- data.frame(
      lambda = lambda,
      eta0 = eta_fit$eta0,
      null_log_marginal = eta_fit$log_marginal,
      eta0_boundary = eta_fit$boundary,
      susie_elbo = susie_elbo,
      converged = isTRUE(fit$converged)
    )
    if (verbose) {
      message(sprintf(
        "null-R: lambda=%g eta0=%.5g SuSiE ELBO=%.6f",
        lambda, eta_fit$eta0, susie_elbo
      ))
    }
  }
  profile <- do.call(rbind, profile)
  best <- which.max(profile$susie_elbo)
  result <- candidate_fits[[best]]
  result$method <- "Null posterior-mean R"
  result$selected_lambda <- profile$lambda[best]
  result$raw_eta0 <- profile$eta0[best]
  result$eta0 <- profile$eta0[best]
  result$eta0_used <- profile$eta0[best]
  result$R_hat <- (
    result$eta0 * ((1 - result$selected_lambda) * R0 +
                     result$selected_lambda * diag(J)) + N * tcrossprod(v)
  ) / (result$eta0 + 1)
  result$profile <- profile
  class(result) <- unique(c("null_estimate_R", class(result)))
  result
}

# -------------------------------------------------------------------------
# 3. SuSiE-IW simulation with two weak-LD causal effects in a 7:5 ratio
# -------------------------------------------------------------------------

choose_weak_ld_variants <- function(R) {
  candidates <- unique(as.integer(round(seq(
    max(2, 0.08 * nrow(R)), min(nrow(R) - 1, 0.92 * nrow(R)),
    length.out = min(250L, nrow(R) - 2L)
  ))))
  minimum_separation <- max(8L, floor(nrow(R) / 12))
  first <- candidates[which.min(abs(candidates - round(nrow(R) / 4)))]
  available <- candidates[abs(candidates - first) >= minimum_separation]
  second <- available[which.min(abs(R[first, available]))]
  c(first, second)
}

simulate_dataset <- function(true_eta0, replication) {
  simulation_seed <- seed * 10000000L + as.integer(true_eta0) * 100L +
    replication
  set.seed(simulation_seed)
  precision <- stats::rWishart(
    1L, df = true_eta0 + J + 1L,
    Sigma = Rbar_true_inverse / true_eta0
  )[, , 1L]
  R <- stats::cov2cor(chol2inv(chol(precision)))
  R <- (R + t(R)) / 2
  diag(R) <- 1

  causal <- choose_weak_ld_variants(R)
  beta_scale <- sqrt(
    h2 / drop(crossprod(
      effect_weights, R[causal, causal, drop = FALSE] %*% effect_weights
    ))
  )
  beta <- numeric(J)
  beta[causal] <- beta_scale * effect_weights
  noise <- as.numeric(t(chol(R)) %*% stats::rnorm(J)) / sqrt(N)
  v <- as.numeric(R %*% beta + noise)
  list(
    v = v, R = R, beta = beta, causal = causal,
    causal_LD = R[causal[1L], causal[2L]],
    expected_z = sqrt(N) * as.numeric(R[causal, , drop = FALSE] %*% beta)
  )
}

evaluate_fit <- function(fit, causal) {
  sets <- susie_get_reported_cs(fit)
  covered <- vapply(causal, function(j) {
    any(vapply(sets, function(set) j %in% set, logical(1L)))
  }, logical(1L))
  true_sets <- sum(vapply(sets, function(set) any(set %in% causal), logical(1L)))
  data.frame(
    estimated_eta0 = if (!is.null(fit$raw_eta0)) {
      fit$raw_eta0
    } else {
      fit$selected_raw_eta0
    },
    eta0_used = if (!is.null(fit$eta0_used)) fit$eta0_used else fit$eta0,
    selected_lambda = fit$selected_lambda,
    reported_L = length(sets),
    true_sets = true_sets,
    false_sets = length(sets) - true_sets,
    causal_power = mean(covered),
    all_causal_covered = all(covered),
    fallback = isTRUE(fit$postprocess$fallback_used),
    converged = isTRUE(fit$converged)
  )
}


# -------------------------------------------------------------------------
# User settings
# -------------------------------------------------------------------------

seed <- 7L
J <- 500L
true_eta0 <- 200
replication <- 1L

N0 <- 2000L
coalescent_sample_size <- 2000L
N <- 200000L
h2 <- 1e-3
true_lambda <- 1e-3
true_L <- 2L
effect_weights <- c(7, 5)
maximum_fitted_L <- 5L
lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)
collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

# -------------------------------------------------------------------------
# Generate one coalescent SuSiE-IW dataset
# -------------------------------------------------------------------------

message("Generating the coalescent reference LD matrix")
R0 <- simulate_coalescent_R0(
  J = J, N0 = N0, second_population_size = coalescent_sample_size,
  seed = seed
)
Rbar_true <- (1 - true_lambda) * R0 + true_lambda * diag(J)
Rbar_true_inverse <- chol2inv(chol(Rbar_true))
simulation <- simulate_dataset(true_eta0, replication)
v <- simulation$v
R <- simulation$R
causal <- simulation$causal

# -------------------------------------------------------------------------
# Fit the two SuSiE-RSS references and the three comparison methods
# -------------------------------------------------------------------------

fits <- list(
  susie_in_sample = fit_susie_rss(
    v = v, R = R, N = N, L = maximum_fitted_L
  ),
  susie_reference = fit_susie_rss(
    v = v, R = R0, N = N, L = maximum_fitted_L
  ),
  null_estimate_R = fit_null_estimate_R(
    v = v, R0 = R0, N = N, L = maximum_fitted_L,
    lambda_grid = lambda_grid, verbose = FALSE
  ),
  collapsed_R = susie_iw(
    v = v, R0 = R0, N = N, L = maximum_fitted_L,
    lambda_grid = lambda_grid,
    eta0_multiplier = collapsed_eta0_multiplier,
    scaled_prior_variance = 0.2, estimate_prior_variance = TRUE,
    coverage = 0.95, min_abs_corr = 0.5,
    ser_fallback = TRUE, partition_ser = FALSE,
    max_iter = 200L, tol = 1e-4, verbose = FALSE
  ),
  AIW_N = suppressMessages(susie_aiw(
    v = v, R0 = R0, N = N, L = maximum_fitted_L,
    approximation = "N", lambda_grid = lambda_grid,
    eta0_multiplier = aiwn_eta0_multiplier,
    scaled_prior_variance = 0.2, estimate_prior_variance = TRUE,
    coverage = 0.95, min_abs_corr = 0.5,
    ser_fallback = TRUE, partition_ser = FALSE,
    max_iter = 500L, tol = 1e-3, verbose = FALSE
  ))
)

method_names <- c(
  "SuSiE-RSS: in-sample R", "SuSiE-RSS: reference R0",
  "Null posterior-mean R", "collapsed-R", "AIW-N"
)
summary_table <- do.call(rbind, lapply(seq_along(fits), function(i) {
  row <- summarize_iw_fit(fits[[i]], causal = causal)
  row$method <- method_names[i]
  row
}))
rownames(summary_table) <- NULL

cat("\nSimulation\n")
print(data.frame(
  J = J, true_eta0 = true_eta0, replication = replication,
  causal_variants = paste(causal, collapse = ","),
  causal_LD = simulation$causal_LD,
  expected_abs_z = paste(round(abs(simulation$expected_z), 3), collapse = ","),
  row.names = NULL
), row.names = FALSE)

cat("\nFits\n")
print(summary_table, row.names = FALSE, digits = 5)

cat("\nSelected mismatch parameters\n")
selected_parameters <- data.frame(
  method = c("Null posterior-mean R", "collapsed-R", "AIW-N"),
  lambda = c(
    fits$null_estimate_R$selected_lambda,
    fits$collapsed_R$selected_lambda,
    fits$AIW_N$selected_lambda
  ),
  raw_eta0 = c(
    fits$null_estimate_R$raw_eta0,
    fits$collapsed_R$raw_eta0,
    fits$AIW_N$selected_raw_eta0
  ),
  eta0_used = c(
    fits$null_estimate_R$eta0,
    fits$collapsed_R$eta0,
    fits$AIW_N$selected_eta0
  ),
  row.names = NULL
)
print(selected_parameters, row.names = FALSE, digits = 6)

cat("\nNull posterior-mean R profile across lambda\n")
print(fits$null_estimate_R$profile, row.names = FALSE, digits = 6)

# -------------------------------------------------------------------------
# PIP comparison in the project's manuscript style
# -------------------------------------------------------------------------

plot_items <- list(
  "SuSiE-RSS: in-sample R" = list(
    object = fits$susie_in_sample, R = R
  ),
  "SuSiE-RSS: reference R0" = list(
    object = fits$susie_reference, R = R0
  ),
  "Null posterior-mean R" = list(
    object = fits$null_estimate_R, R = fits$null_estimate_R$R_hat
  ),
  "Collapsed-R" = list(
    object = fits$collapsed_R, R = fits$collapsed_R$selected_Rbar
  ),
  "AIW-N" = list(
    object = fits$AIW_N, R = fits$AIW_N$selected_Rbar
  )
)

draw_pip_comparison <- function() {
  plot_iw_comparison(
    plot_items, causal = causal, ncol = 2L,
    main = sprintf(
      "Coalescent SuSiE-IW: J=%d, true eta0=%s, replication=%d",
      J, format(true_eta0, big.mark = ",", scientific = FALSE), replication
    )
  )
}

output_stem <- file.path(
  "experiments", "04-null-estimate-R-demonstration-PIP"
)
grDevices::png(
  paste0(output_stem, ".png"), width = 1800, height = 2100, res = 180
)
draw_pip_comparison()
grDevices::dev.off()
grDevices::pdf(paste0(output_stem, ".pdf"), width = 10, height = 11.7)
draw_pip_comparison()
grDevices::dev.off()

draw_pip_comparison()
```
