---
title: "Comparison with the susieR empirical-Bayes mismatch correction"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  run_analysis: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 11.7
)
```

This page compares the project's mismatch-aware methods with the empirical-
Bayes LD mismatch correction in `susieR::susie_rss()`. The simulation matches
the [coalescent SuSiE-IW simulation](02-coalescent-IW-simulation.html):
coalescent reference LD is followed by a correlation-standardized
inverse-Wishart draw of the in-sample LD matrix and SuSiE-IW summary statistics
with two or three causal effects.

The susieR fit follows the official [LD mismatch
workflow](https://stephenslab.github.io/susieR/articles/rss_mismatch.html#empirical-bayes-mismatch-correction):

```r
susie_rss(..., R_finite = N0, R_mismatch = "eb")
```

`R_finite` accounts for estimation noise from the finite reference panel, and
`R_mismatch = "eb"` estimates an additional region-wide mismatch component.
The analysis reports `R_finite_diagnostics` because the corrected result should
not be treated as automatically reliable.

The five comparisons are in-sample SuSiE-RSS, reference-LD SuSiE-RSS,
reference-LD SuSiE-RSS with finite-panel and empirical-Bayes mismatch
correction, collapsed-$R$, and AIW-N. This page requires a susieR version that
provides `R_finite`, `R_mismatch`, and `susie_rss_control` (version 0.16.6 or
newer at the time this page was written).

Set `run_analysis: false` in the YAML header to skip the computationally
nontrivial coalescent simulation and fits during a workflowr build.

## Analysis

```{r analysis, eval=params$run_analysis}

find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  repeat {
    required <- c(
      file.path(path, "code", "01-collapsed-R.R"),
      file.path(path, "code", "02-AIW-and-AIW-N.R"),
      file.path(path, "code", "03-SER-fallback-and-plot.R")
    )
    if (all(file.exists(required))) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find the susie-IW project root")
    }
    path <- parent
  }
}

project_root <- find_project_root()
source(file.path(project_root, "code", "01-collapsed-R.R"))
source(file.path(project_root, "code", "02-AIW-and-AIW-N.R"))
source(file.path(project_root, "code", "03-SER-fallback-and-plot.R"))

if (!requireNamespace("susieR", quietly = TRUE)) {
  stop("Install the current susieR package before running this comparison")
}
required_susie_rss_arguments <- c("R_finite", "R_mismatch", "control")
missing_susie_rss_arguments <- setdiff(
  required_susie_rss_arguments,
  names(formals(susieR::susie_rss))
)
if (length(missing_susie_rss_arguments)) {
  stop(
    "Installed susieR ", as.character(utils::packageVersion("susieR")),
    " does not support the current mismatch interface (missing: ",
    paste(missing_susie_rss_arguments, collapse = ", "), "). ",
    "Install susieR >= 0.16.6, for example with ",
    "remotes::install_github('stephenslab/susieR')."
  )
}

# -------------------------------------------------------------------------
# Simulation and fitting settings. Edit this block to select the experiment.
# -------------------------------------------------------------------------

seed <- 1L
J <- 1000L
N0 <- 2000L                       # reference-panel sample size
coalescent_sample_size <- 2000L   # second population used for common-SNP QC
N <- 200000L                      # GWAS sample size
true_eta0 <- 100
h2 <- 1e-3
true_lambda <- 1e-3
true_L_values <- c(2L, 3L)
maximum_fitted_L <- 5L

lambda_grid <- c(1e-5, 1e-4, 1e-3, 2e-3, 4e-3, 6e-3, 8e-3, 1e-2)
collapsed_eta0_multiplier <- 1.677
aiwn_eta0_multiplier <- 1.424

if (J <= max(true_L_values) || maximum_fitted_L < max(true_L_values) ||
    maximum_fitted_L >= J) {
  stop("Require max(true_L_values) <= maximum_fitted_L < J")
}

output_root <- file.path(project_root, "experiments")
figure_directory <- file.path(output_root, "figures")
dir.create(figure_directory, recursive = TRUE, showWarnings = FALSE)
output_prefix <- file.path(output_root, "05-compare-susie-rss")
figure_prefix <- file.path(figure_directory, "05-compare-susie-rss")

# -------------------------------------------------------------------------
# Generate empirical reference LD from the same human coalescent model used in
# analysis/02-coalescent-IW-simulation.Rmd.
# -------------------------------------------------------------------------

simulate_coalescent_R0 <- function(
    J, N0, second_population_size = N0, seed = 1L,
    chromosome = "chr22", start = 20e6, end = 21e6,
    split_time = 10L) {
  if (!requireNamespace("reticulate", quietly = TRUE)) {
    stop("Install the R package reticulate")
  }
  if (!nzchar(Sys.getenv("RETICULATE_PYTHON"))) {
    python <- Sys.which("python3")
    if (nzchar(python)) {
      Sys.setenv(RETICULATE_PYTHON = python)
      reticulate::use_python(python, required = FALSE)
    }
  }
  stdpopsim <- reticulate::import("stdpopsim")
  msprime <- reticulate::import("msprime")

  species <- stdpopsim$get_species("HomSap")
  contig <- species$get_contig(
    chromosome, left = as.integer(start), right = as.integer(end)
  )
  demography <- msprime$Demography()
  demography$add_population(name = "ANC", initial_size = 20000L)
  demography$add_population(name = "POP1", initial_size = 100000L)
  demography$add_population(name = "POP2", initial_size = 100000L)
  demography$add_population_split(
    time = as.integer(split_time),
    derived = list("POP1", "POP2"), ancestral = "ANC"
  )

  models <- list(
    msprime$DiscreteTimeWrightFisher(duration = 100L),
    msprime$StandardCoalescent()
  )
  tree_sequence <- msprime$sim_ancestry(
    samples = reticulate::dict(
      POP1 = as.integer(second_population_size),
      POP2 = as.integer(N0)
    ),
    demography = demography,
    recombination_rate = contig$recombination_map,
    model = models,
    random_seed = as.integer(seed)
  )
  tree_sequence <- msprime$sim_mutations(
    tree_sequence, rate = 1.29e-8, random_seed = 1L
  )

  population_ids <- list()
  for (population in reticulate::iterate(tree_sequence$populations())) {
    population_ids[[population$metadata$name]] <- population$id
  }
  genotype_haploid <- tree_sequence$genotype_matrix()
  to_diploid <- function(sample_ids) {
    columns <- as.integer(sample_ids) + 1L
    haplotypes <- genotype_haploid[, columns, drop = FALSE]
    genotypes <- haplotypes[, seq(1L, ncol(haplotypes), by = 2L)] +
      haplotypes[, seq(2L, ncol(haplotypes), by = 2L)]
    t(genotypes)
  }
  genotype_1 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP1)
  ))
  genotype_0 <- to_diploid(tree_sequence$samples(
    population = as.integer(population_ids$POP2)
  ))

  frequency_1 <- colMeans(genotype_1) / 2
  frequency_0 <- colMeans(genotype_0) / 2
  common <- frequency_1 > 0.05 & frequency_1 < 0.95 &
    frequency_0 > 0.05 & frequency_0 < 0.95
  genotype_0 <- genotype_0[, common, drop = FALSE]
  if (ncol(genotype_0) < J) {
    stop("The coalescent region contains fewer than J common variants")
  }

  set.seed(seed)
  first_column <- if (ncol(genotype_0) == J) {
    1L
  } else {
    sample.int(ncol(genotype_0) - J + 1L, 1L)
  }
  columns <- first_column + seq_len(J) - 1L
  R0 <- stats::cor(genotype_0[, columns, drop = FALSE])
  R0 <- (R0 + t(R0)) / 2
  diag(R0) <- 1
  R0
}

message("Generating coalescent reference LD")
R0 <- simulate_coalescent_R0(
  J = J, N0 = N0, second_population_size = coalescent_sample_size,
  seed = seed
)

# -------------------------------------------------------------------------
# Draw in-sample LD under the correlation-standardized SuSiE-IW simulation.
# -------------------------------------------------------------------------

Rbar_true <- (1 - true_lambda) * R0 + true_lambda * diag(J)
set.seed(2000000L + 10000L * seed + J + as.integer(true_eta0))
precision <- stats::rWishart(
  1L,
  df = true_eta0 + J + 1L,
  Sigma = solve(true_eta0 * Rbar_true)
)[, , 1L]
R <- stats::cov2cor(solve(precision))
R <- (R + t(R)) / 2
diag(R) <- 1

choose_nested_weak_ld_variants <- function(R) {
  candidates <- unique(as.integer(round(seq(
    max(2, 0.08 * nrow(R)), min(nrow(R) - 1, 0.92 * nrow(R)),
    length.out = min(250L, nrow(R) - 2L)
  ))))
  minimum_separation <- max(8L, floor(nrow(R) / 12))
  first <- candidates[which.min(abs(candidates - round(nrow(R) / 4)))]
  available <- candidates[abs(candidates - first) >= minimum_separation]
  second <- available[which.min(abs(R[first, available]))]
  available <- candidates[
    abs(candidates - first) >= minimum_separation &
      abs(candidates - second) >= minimum_separation
  ]
  third <- available[which.min(vapply(available, function(j) {
    max(abs(R[j, c(first, second)]))
  }, numeric(1L)))]
  c(first, second, third)
}

causal_candidates <- choose_nested_weak_ld_variants(R)
set.seed(2500000L + 10000L * seed + J + as.integer(true_eta0))
noise <- as.numeric(t(chol(R)) %*% stats::rnorm(J)) / sqrt(N)

simulate_architecture <- function(true_L) {
  stopifnot(true_L %in% true_L_values)
  causal <- causal_candidates[seq_len(true_L)]
  weights <- c(7, 5, 3)[seq_len(true_L)]
  effect_scale <- sqrt(
    h2 / drop(crossprod(
      weights, R[causal, causal, drop = FALSE] %*% weights
    ))
  )
  beta <- numeric(J)
  beta[causal] <- effect_scale * weights
  v <- as.numeric(R %*% beta + noise)
  list(
    true_L = true_L,
    causal = causal,
    beta = beta,
    v = v,
    causal_LD = R[causal, causal, drop = FALSE],
    expected_z = sqrt(N) * as.numeric(
      R[causal, , drop = FALSE] %*% beta
    )
  )
}

simulations <- setNames(
  lapply(true_L_values, simulate_architecture),
  paste0("L", true_L_values)
)

# -------------------------------------------------------------------------
# Fit the four existing comparison methods plus susie_rss with finite-panel
# uncertainty and empirical-Bayes mismatch correction.
# -------------------------------------------------------------------------

fit_susie_rss_eb_mismatch <- function(v, R0, N, N0, L) {
  z <- iw_z_from_v(v, N)
  susieR::susie_rss(
    z = z,
    R = R0,
    n = N,
    L = as.integer(L),
    scaled_prior_variance = 0.2,
    estimate_prior_variance = TRUE,
    coverage = 0.95,
    min_abs_corr = 0.5,
    max_iter = 500L,
    tol = 1e-3,
    R_finite = N0,
    R_mismatch = "eb",
    track_fit = TRUE
  )
}

fit_architecture <- function(simulation) {
  v <- simulation$v
  list(
    "SuSiE-RSS: in-sample R" = fit_susie_rss(
      v = v, R = R, N = N, L = maximum_fitted_L
    ),
    "SuSiE-RSS: reference R0" = fit_susie_rss(
      v = v, R = R0, N = N, L = maximum_fitted_L
    ),
    "SuSiE-RSS: finite R0 + EB mismatch" = fit_susie_rss_eb_mismatch(
      v = v, R0 = R0, N = N, N0 = N0, L = maximum_fitted_L
    ),
    "Collapsed-R" = susie_iw(
      v = v, R0 = R0, N = N, L = maximum_fitted_L,
      lambda_grid = lambda_grid,
      eta0_multiplier = collapsed_eta0_multiplier,
      scaled_prior_variance = 0.2,
      estimate_prior_variance = TRUE,
      coverage = 0.95, min_abs_corr = 0.5,
      ser_fallback = TRUE,
      max_iter = 200L, tol = 1e-4, verbose = FALSE
    ),
    "AIW-N" = suppressMessages(susie_aiw(
      v = v, R0 = R0, N = N, L = maximum_fitted_L,
      approximation = "N",
      lambda_grid = lambda_grid,
      eta0_multiplier = aiwn_eta0_multiplier,
      scaled_prior_variance = 0.2,
      estimate_prior_variance = TRUE,
      coverage = 0.95, min_abs_corr = 0.5,
      ser_fallback = TRUE,
      max_iter = 500L, tol = 1e-3, verbose = FALSE
    ))
  )
}

fits <- lapply(names(simulations), function(label) {
  message("Fitting architecture ", label)
  fit_architecture(simulations[[label]])
})
names(fits) <- names(simulations)

# -------------------------------------------------------------------------
# Numerical summaries, including the reliability diagnostics emphasized in the
# official susieR vignette.
# -------------------------------------------------------------------------

summarize_architecture <- function(label, simulation, fitted_methods) {
  table <- do.call(rbind, Map(function(method, fit) {
    row <- summarize_iw_fit(fit, causal = simulation$causal)
    row$architecture <- label
    row$method <- method
    row
  }, names(fitted_methods), fitted_methods))
  rownames(table) <- NULL
  table[, c(
    "architecture", "method",
    setdiff(names(table), c("architecture", "method"))
  )]
}

summary_table <- do.call(rbind, lapply(names(fits), function(label) {
  summarize_architecture(label, simulations[[label]], fits[[label]])
}))
rownames(summary_table) <- NULL

numeric_scalar <- function(value) {
  if (is.null(value) || !length(value)) return(NA_real_)
  as.numeric(value[1L])
}

logical_scalar <- function(value) {
  if (is.null(value) || !length(value)) return(NA)
  as.logical(value[1L])
}

character_scalar <- function(value) {
  if (is.null(value) || !length(value)) return(NA_character_)
  as.character(value[1L])
}

eb_diagnostic_table <- do.call(rbind, lapply(names(fits), function(label) {
  fit <- fits[[label]][["SuSiE-RSS: finite R0 + EB mismatch"]]
  diagnostics <- fit$R_finite_diagnostics
  data.frame(
    architecture = label,
    susieR_version = as.character(utils::packageVersion("susieR")),
    R_finite = numeric_scalar(diagnostics$B),
    R_mismatch = character_scalar(diagnostics$R_mismatch),
    mismatch_estimator = character_scalar(fit$R_mismatch_method),
    lambda_bias = numeric_scalar(fit$lambda_bias),
    corrected_reference_size = numeric_scalar(fit$B_corrected),
    Q_art = numeric_scalar(diagnostics$Q_art),
    residual_artifact_flag = logical_scalar(diagnostics$artifact_flag),
    CS_sensitivity_flag = logical_scalar(diagnostics$R_sensitivity_flag),
    overall_reliability_flag = logical_scalar(
      diagnostics$R_reliability_flag
    ),
    convergence_reason = character_scalar(fit$convergence_reason),
    converged = logical_scalar(fit$converged),
    iterations = numeric_scalar(fit$niter),
    stringsAsFactors = FALSE
  )
}))
rownames(eb_diagnostic_table) <- NULL

generating_table <- do.call(rbind, lapply(names(simulations), function(label) {
  simulation <- simulations[[label]]
  data.frame(
    architecture = label,
    true_L = simulation$true_L,
    causal = paste(simulation$causal, collapse = ","),
    expected_abs_z = paste(round(abs(simulation$expected_z), 2), collapse = ","),
    max_abs_causal_LD = max(abs(
      simulation$causal_LD[upper.tri(simulation$causal_LD)]
    )),
    stringsAsFactors = FALSE
  )
}))
rownames(generating_table) <- NULL

utils::write.csv(
  summary_table, paste0(output_prefix, "-summary.csv"), row.names = FALSE
)
utils::write.csv(
  eb_diagnostic_table,
  paste0(output_prefix, "-eb-diagnostics.csv"),
  row.names = FALSE
)
utils::write.csv(
  generating_table,
  paste0(output_prefix, "-generating-settings.csv"),
  row.names = FALSE
)

cat("\nMethod comparison\n")
print(summary_table, row.names = FALSE, digits = 4)
cat("\nsusie_rss empirical-Bayes mismatch diagnostics\n")
print(eb_diagnostic_table, row.names = FALSE, digits = 5)
cat("\nGenerating architectures\n")
print(generating_table, row.names = FALSE, digits = 4)

# -------------------------------------------------------------------------
# PIP and credible-set comparisons.
# -------------------------------------------------------------------------

plot_items_for_architecture <- function(label) {
  current_fits <- fits[[label]]
  list(
    "SuSiE-RSS: in-sample R" = list(
      object = current_fits[["SuSiE-RSS: in-sample R"]], R = R
    ),
    "SuSiE-RSS: reference R0" = list(
      object = current_fits[["SuSiE-RSS: reference R0"]], R = R0
    ),
    "SuSiE-RSS: finite R0 + EB mismatch" = list(
      object = current_fits[["SuSiE-RSS: finite R0 + EB mismatch"]], R = R0
    ),
    "Collapsed-R" = list(
      object = current_fits[["Collapsed-R"]],
      R = current_fits[["Collapsed-R"]]$selected_Rbar
    ),
    "AIW-N" = list(
      object = current_fits[["AIW-N"]],
      R = current_fits[["AIW-N"]]$selected_Rbar
    )
  )
}

draw_architecture <- function(label) {
  plot_iw_comparison(
    plot_items_for_architecture(label),
    causal = simulations[[label]]$causal,
    ncol = 2L,
    main = paste0("Coalescent SuSiE-IW comparison: ", label)
  )
}

for (label in names(fits)) {
  filename <- paste0(figure_prefix, "-", label)
  grDevices::png(
    paste0(filename, ".png"), width = 1800, height = 2100, res = 180
  )
  draw_architecture(label)
  grDevices::dev.off()
  grDevices::pdf(paste0(filename, ".pdf"), width = 10, height = 11.7)
  draw_architecture(label)
  grDevices::dev.off()
}

# Draw the two comparisons again so knitr captures them in the rendered page.
for (label in names(fits)) draw_architecture(label)

message("Wrote comparison outputs with prefix: ", output_prefix)
```
