---
title: "Correlation-standardized collapsed-R heuristic"
author: "Dat Do"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
params:
  rerun_analysis: false
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Question

The exact collapsed-R update supplies

\[
\widehat R
= E_q(\omega^{-1})\{\eta_0\bar R+Nvv^\top\}
\]

to the IBSS update. Its diagonal is generally not one. This page examines the
heuristic replacement

\[
\widetilde R=\operatorname{cov2cor}(\widehat R).
\]

Only the update of the variational distribution of the regression effects is
changed. The updates of (q(\omega)), \(\eta_0\), and the effect-specific prior
variances, and the displayed collapsed-R ELBO, are still calculated under the
original model. Consequently, the displayed ELBO is a diagnostic and is not
guaranteed to increase.

The positive scalar (E_q(\omega^{-1})) cancels from `cov2cor`, so the working
matrix is equivalently

\[
\widetilde R=\operatorname{cov2cor}
\{\eta_0\bar R+Nvv^\top\}.
\]

## Simulation and fits

We generate reference LD (R_0) from a human coalescent simulation, draw the
in-sample LD (R) from the SuSiE-IW model, and simulate two weak-LD effects in
the ratio (7:5). To focus on the new update, the fitted value of \(\lambda\)
is fixed at its generating value. We compare the exact collapsed-R algorithm,
the heuristic with \(\eta_0\) fixed at its true value, and the operational
heuristic that learns \(\eta_0\). The heuristic fits are deliberately continued
for 200 iterations instead of stopping at the first apparent plateau.

```{r run-or-load}
result_file <- file.path(
  "experiments", "heuristic-collapsed-R-elbo-study.rds"
)

if (isTRUE(params$rerun_analysis) || !file.exists(result_file)) {
  source(file.path("experiments", "simulation-iw-collapsed-R.R"))
  source(file.path("code", "02-AIW-and-AIW-N.R"))
  source(file.path("code", "03-SER-fallback-and-plot.R"))
  source(file.path("code", "04-heuristic-collapsed-R.R"))

  seed <- 17L
  J <- 100L
  N0 <- 1000L
  N <- 200000L
  h2 <- 1e-3
  true_lambda <- 1e-3
  true_eta0 <- 100
  true_L <- 2L
  fitted_L <- 5L
  long_run_iterations <- 200L

  R0 <- simulate_coalescent_reference_ld(
    coalescent_sample_size = N0,
    reference_sample_size = N0,
    number_of_variants = J,
    seed = seed,
    heritability = h2
  )
  Rbar_true <- make_stabilized_reference_ld(R0, true_lambda)

  set.seed(1000000L + seed)
  R <- standardize_correlation(sample_inverse_wishart(
    degrees_of_freedom = true_eta0 + J + 1,
    scale_matrix = true_eta0 * Rbar_true
  ))

  candidates <- unique(as.integer(round(seq(
    max(2, 0.1 * J), min(J - 1, 0.9 * J),
    length.out = min(80L, J - 2L)
  ))))
  first <- candidates[which.min(abs(candidates - round(J / 3)))]
  available <- candidates[abs(candidates - first) >= floor(J / 4)]
  second <- available[which.min(abs(R[first, available]))]
  causal <- c(first, second)

  beta <- numeric(J)
  beta[causal] <- c(7, 5)
  beta <- beta * sqrt(h2 / drop(crossprod(beta, R %*% beta)))
  set.seed(2000000L + seed)
  noise <- as.numeric(t(chol(R)) %*% stats::rnorm(J)) / sqrt(N)
  v <- as.numeric(R %*% beta + noise)

  exact_time <- system.time({
    exact <- fit_collapsed_r(
      v = v, R0 = R0, N = N, L = fitted_L,
      lambda_grid = true_lambda,
      eta0 = NULL, eta0_multiplier = 1,
      max_iter = long_run_iterations, tol = 1e-4,
      track_elbo_updates = TRUE
    )
  })
  fixed_time <- system.time({
    heuristic_fixed <- fit_collapsed_r_heuristic(
      v = v, R0 = R0, N = N, L = fitted_L,
      lambda_grid = true_lambda,
      eta0 = true_eta0, eta0_multiplier = 1,
      max_iter = long_run_iterations, tol = 1e-6,
      stop_when_stable = FALSE, track_elbo_updates = TRUE
    )
  })
  learned_time <- system.time({
    heuristic_learned <- fit_collapsed_r_heuristic(
      v = v, R0 = R0, N = N, L = fitted_L,
      lambda_grid = true_lambda,
      eta0 = NULL, eta0_multiplier = 1,
      max_iter = long_run_iterations, tol = 1e-6,
      stop_when_stable = FALSE, track_elbo_updates = TRUE
    )
  })

  study <- list(
    settings = list(
      seed = seed, J = J, N0 = N0, N = N, h2 = h2,
      true_lambda = true_lambda, true_eta0 = true_eta0,
      true_L = true_L, fitted_L = fitted_L,
      long_run_iterations = long_run_iterations,
      causal = causal, causal_LD = R[causal[1L], causal[2L]]
    ),
    data = list(R0 = R0, R = R, Rbar_true = Rbar_true,
                beta = beta, v = v),
    fits = list(
      `Exact collapsed-R` = exact,
      `Heuristic: fixed eta0` = heuristic_fixed,
      `Heuristic: learned eta0` = heuristic_learned
    ),
    elapsed = c(
      `Exact collapsed-R` = unname(exact_time["elapsed"]),
      `Heuristic: fixed eta0` = unname(fixed_time["elapsed"]),
      `Heuristic: learned eta0` = unname(learned_time["elapsed"])
    )
  )
  saveRDS(study, result_file)
} else {
  study <- readRDS(result_file)
}

settings <- study$settings
fits <- study$fits
```

The simulated causal variants are `r paste(settings$causal, collapse = " and ")`;
their in-sample correlation is `r sprintf("%.3f", settings$causal_LD)`.
The dimensions are \(J=`r settings$J`\), \(N_0=`r settings$N0`\), and
\(N=`r format(settings$N, big.mark = ",")`\), with true
\(\eta_0=`r settings$true_eta0`\).

## Diagnostic ELBO

```{r prepare-results}
fit_core <- function(object) object$fit

trajectory <- do.call(rbind, lapply(names(fits), function(method) {
  values <- fit_core(fits[[method]])$elbo
  data.frame(
    method = method,
    iteration = seq_along(values),
    elbo = values,
    centered_elbo = values - max(values),
    elbo_change = c(NA_real_, diff(values))
  )
}))
trajectory$method <- factor(trajectory$method, levels = names(fits))

state <- do.call(rbind, lapply(names(fits)[-1L], function(method) {
  value <- fit_core(fits[[method]])$state_trace
  value$method <- method
  value
}))
state$method <- factor(state$method, levels = names(fits)[-1L])
```

```{r elbo-trajectory, fig.cap="Exact collapsed-R ELBO evaluated along each algorithm's iterates. Centering is performed separately within each fit. In this example the fixed-eta0 heuristic approaches a plateau, whereas the learned-eta0 heuristic moves away from its early ELBO maximum."}
library(ggplot2)

ggplot(trajectory, aes(iteration, centered_elbo, color = method)) +
  geom_hline(yintercept = 0, color = "grey75", linewidth = 0.35) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~method, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("#1F78B4", "#33A02C", "#E31A1C")) +
  labs(x = "Iteration", y = "ELBO minus maximum within fit") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", strip.background = element_blank(),
        panel.grid.minor = element_blank())
```

Because an inactive effect-specific prior variance can continue shrinking after
the PIPs have settled, the next plot separates changes in the posterior
allocation, \(\eta_0\), and prior variance.

```{r state-change, fig.cap="Changes in the state of the two long-run heuristic fits. Values at zero are replaced only for display on the logarithmic scale."}
state_long <- rbind(
  data.frame(method = state$method, iteration = state$iteration,
             component = "alpha", change = state$alpha_change),
  data.frame(method = state$method, iteration = state$iteration,
             component = "eta0", change = state$eta0_log_change),
  data.frame(method = state$method, iteration = state$iteration,
             component = "prior variance", change = state$sigma2_log_change)
)
state_long$change_for_plot <- pmax(state_long$change, 1e-12)

ggplot(state_long, aes(iteration, change_for_plot, color = component)) +
  geom_line(linewidth = 0.65) +
  facet_wrap(~method, ncol = 1) +
  scale_y_log10() +
  scale_color_manual(values = c(alpha = "#1F78B4", eta0 = "#E31A1C",
                                `prior variance` = "#6A3D9A")) +
  labs(x = "Iteration", y = "Absolute change", color = NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.background = element_blank(),
        panel.grid.minor = element_blank())
```

## Endpoint summary

```{r endpoint-summary}
summarize_fit <- function(method, object) {
  core <- fit_core(object)
  changes <- diff(core$elbo)
  tail_values <- tail(core$elbo, min(20L, length(core$elbo)))
  causal_pip <- core$pip[settings$causal]
  data.frame(
    method = method,
    iterations = length(core$elbo),
    eta0 = core$nu0,
    negative_ELBO_steps = sum(changes < -1e-10),
    largest_ELBO_drop = if (length(changes)) min(0, changes) else NA_real_,
    ELBO_range_last_20 = diff(range(tail_values)),
    maximum_causal_PIP = max(causal_pip),
    minimum_causal_PIP = min(causal_pip),
    seconds = unname(study$elapsed[method])
  )
}

endpoint <- do.call(rbind, Map(summarize_fit, names(fits), fits))
rownames(endpoint) <- NULL
knitr::kable(endpoint, digits = 4)
```

```{r pip-comparison, fig.cap="Endpoint PIPs. Red circles identify the two simulated causal variants."}
pip_data <- do.call(rbind, lapply(names(fits), function(method) {
  data.frame(method = method, variant = seq_len(settings$J),
             pip = fit_core(fits[[method]])$pip)
}))
pip_data$method <- factor(pip_data$method, levels = names(fits))

ggplot(pip_data, aes(variant, pip)) +
  geom_point(size = 0.8, color = "black") +
  geom_point(
    data = subset(pip_data, variant %in% settings$causal),
    shape = 1, stroke = 1.1, size = 2.8, color = "firebrick"
  ) +
  facet_wrap(~method, ncol = 1) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Variant index", y = "PIP") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_blank(), panel.grid.minor = element_blank())
```

## Findings

The fixed-\(\eta_0\) heuristic behaves well in this example. Its exact-model
ELBO happens to increase toward a plateau, although monotonicity is not
guaranteed, and the two causal variants receive PIPs of
`r paste(sprintf("%.3f", fit_core(fits[["Heuristic: fixed eta0"]])$pip[settings$causal]), collapse = " and ")`.
The posterior allocations change by only
`r sprintf("%.2g", tail(fit_core(fits[["Heuristic: fixed eta0"]])$state_trace$alpha_change, 1L))`
at iteration 200. An inactive effect-specific prior variance continues to
shrink slowly, which explains why the full state-change diagnostic has not yet
reached the strict numerical tolerance.

The learned-\(\eta_0\) heuristic does **not** behave well. It selects
\(\widehat\eta_0=`r signif(fit_core(fits[["Heuristic: learned eta0"]])$nu0, 4)`\),
the lower edge of the search interval, and has
`r sum(diff(fit_core(fits[["Heuristic: learned eta0"]])$elbo) < -1e-10)`
decreasing ELBO steps. Its allocation is still changing after 200 iterations
and it retains only one of the two causal signals while assigning very large
PIPs to several noncausal variants.

This failure has a natural feedback interpretation. As \(\eta_0\) becomes
small, \(\eta_0\bar R+Nvv^\top\) becomes dominated by the rank-one term
\(Nvv^\top\). Correlation standardization then converts that term into an
extreme working correlation pattern aligned with the observed marginal
associations. The resulting effect allocation can make an even smaller
\(\eta_0\) attractive under the original-model profiling step. The update of
\(q(b)\) and the update of \(\eta_0\) are therefore not optimizing a common
objective, and the feedback need not settle at a useful fixed point.

The exact collapsed-R ELBO should be retained as a diagnostic, but it cannot
serve as a convergence certificate for this heuristic. More importantly, the
jointly learned version should **not** replace the exact collapsed-R algorithm
in primary analyses without an additional stabilizing rule. The fixed-\(\eta_0\)
version remains useful for controlled experiments, including the proposed
choice \(\eta_0=N_0\), but its empirical behavior must be evaluated separately.
