---
title: "Exploring ideas for dealing with LD mismatch in SuSiE"
author: Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

This vignette builds on my
[JSM tutorial](https://stephenslab.github.io/finemap/jsm_tutorial.html).
It contains some additional explorations on SuSiE and the effects of
errors in the LD matrix on errors in the fine-mapping results.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

For this tutorial, I'm using a lightly modified version of susieR
available in the "output_residuals" branch. I would recommend using this
version to reproduce the same results I got.

```{r install-susie, eval=FALSE}
library(remotes)
install_github("stephenslab/susieR@output_residuals")
```

Load all the packages needed for this analysis. Note the version of
susieR that I am using.

```{r load-pkgs}
library(ggplot2)
library(cowplot)
library(susieR)
packageVersion("susieR")
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

The data set contains genotype and quantitative trait data for $n
= 574$ individuals and $p = 750$ genetic markers (SNPs).

```{r load-data}
load("jsm_tutorial_data.RData")
n <- nrow(X)
dim(X)
```

This is a function we will use in the code chunks below to visualize
the results of the SuSiE fine-mapping analysis.

```{r susie-plot-better}
pip_plot <- function (fit, causal_snps = NULL, xlab = "") {
  cs_colors <- c("dodgerblue","limegreen","darkorange")
  pip <- fit$pip
  n   <- length(pip)
  dat <- data.frame(pos = 1:n,pip = pip,CS = as.character(NA))
  num_cs <- length(fit$sets$cs)
  for (i in num_cs:1) {
    j <- fit$sets$cs[[i]]
	dat[j,"CS"] <- paste0("CS",i)
  }
  dat_cs <- subset(dat,!is.na(CS))
  out <- ggplot(dat,aes(x = pos,y = pip)) +
    geom_point() +
	geom_point(data = dat_cs,shape = 1,size = 3,
	           mapping = aes(color = CS)) +
	scale_x_continuous(breaks = NULL) +
	scale_color_manual(values = cs_colors) + 
	labs(x = xlab,y = "PIP") +
    theme_cowplot(font_size = 12)
  if (!is.null(causal_snps)) {
	dat_causal <- dat[causal_snps,]
    out <- out + geom_point(data = dat_causal,color = "red")
  }
  return(out)
}
```

When we use the "in-sample" LD to fit the SuSiE model, SuSiE correctly
predicts that there are two causal SNPs (the third CS is very diffuse
and will be shrunk to zero we let the prior adapt to the data):

```{r susie-rss, fig.height=1.75, fig.width=5.25}
fit <- susie_rss(zhat,R,n = n,L = 3,estimate_prior_variance = FALSE,
                 tol = 1e-8,min_abs_corr = 0)
causal_snps <- which(b_true != 0)
pip_plot(fit,causal_snps) +
  ggtitle("in-sample LD")
```

On the other hand, when we use the "out-of-sample" LD, SuSiE produces
a "ghost CS" that does not contain a true causal SNP:

```{r susie-rss-out, fig.height=1.75, fig.width=5.25}
fit_out <- susie_rss(zhat,R_out,n = n,L = 3,estimate_prior_method = "EM",
                     tol = 1e-8,min_abs_corr = 0)
pip_plot(fit_out,causal_snps) +
  ggtitle("out-of-sample LD")
```

Examining the residuals gives some insights into the differences
between the SuSiE results with the in-sample and out-of-sample LD
matrices.

```{r residuals-plot}
residuals_plot <- function (fit, z, n, l, cs_color = "red",
                                    title = "") {
  cs  <- fit$sets$cs[[l]]
  dat <- data.frame(z = z,res = fit$XtR[,l]/sqrt(n))
  dat_cs <- dat[cs,]
  zmin <- min(c(dat$z,dat$res))
  zmax <- max(c(dat$z,dat$res))
  return(ggplot(dat,aes(x = z,y = res)) +
         geom_point() +
    	 geom_point(data = dat_cs,shape = 1,size = 3,color = cs_color) +
		 geom_abline(intercept = 0,slope = 1,color = "magenta",
		             linetype = "dotted")  +
	     xlim(zmin,zmax) +
		 ylim(zmin,zmax) +
		 labs(x = "z-score",y = "adjusted z-score",title = title) +
		 theme_cowplot(font_size = 12) +
		 theme(plot.title = element_text(face = "plain",size = 12)))
}
```

In this scatterplot, we compare the z-scores against the "adjusted
z-scores" that were used to fit the third SER (using the in-sample
LD). These adjusted z-scores were obtained by removing posterior mean
effects of the first and second SERs from the z-scores.

```{r residuals-plot-ser-3, fig.height=2.75, fig.width=5}
z <- sqrt(n/(n + zhat^2)) * zhat
p1 <- residuals_plot(fit,z,n,l = 3,title = "in-sample LD")
p2 <- residuals_plot(fit_out,z,n,l = 3,title = "out-of-sample LD")
plot_grid(p1,p2,nrow = 1,ncol = 2)
```

Notice that the use of an out-of-sample LD introduces errors when
adjusting the z-scores, and therefore we end up with "noisier" data.

It is helpful to understand this result in more detail. While it is
true the out-of-sample LD introduces errors, it introduces errors in a
specific way: the errors in the out-of-sample LD matter that matter
most are the correlations with the SNPs in the first two CSs.

```{r residuals-more-1, fig.height=2.5, fig.width=5}
j1 <- 679
j2 <- 129
pdat1 <- data.frame(res_err = fit$XtR[,3] - fit_out$XtR[,3],
                    cor_err = R_out[,j1] - R[,j1])
pdat2 <- data.frame(res_err = fit$XtR[,3] - fit_out$XtR[,3],
                    cor_err = -(R_out[,j2] - R[,j2]))
p1 <- ggplot(pdat1,aes(x = res_err,y = cor_err)) +
  geom_point() +
  labs(x = "error in z-scores",y = "correlation error") +
  theme_cowplot(font_size = 12)
p2 <- ggplot(pdat2,aes(x = res_err,y = cor_err)) +
  geom_point() +
  labs(x = "error in z-scores",y = "correlation error") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,nrow = 1,ncol = 2)
```

Next, I'm testing out Dat's method, which deals with the LD mismatch
in a model-based way.

```{r compute-summary-stats}
n <- nrow(X)
p <- ncol(X)
X <- scale(X,center = TRUE,scale = TRUE)
y <- drop(scale(y,center = TRUE,scale = TRUE))
v <- drop(crossprod(X,y))/n
R_out <- 0.999*R_out + 0.001*diag(p)
```

This function returns a CS given the PIPs:

```{r get-cs}
get_cs <- function (pip, coverage = 0.95) {
  i <- order(pip,decreasing = TRUE)
  n <- which(cumsum(pip[i]) > coverage)[1]
  return(i[1:n])
}
```

When the prior on ${\bf R}$ is highly informative, Dat's method
essentially recovers the same result as the curent `susie_rss()` that
ignores LD mismatch:

```{r aiw-1, fig.height=1.75, fig.width=5.25, results="hide", message=FALSE}
source("code/02-AIW-and-AIW-N.R")
out <- fit_aiwn(v,R_out,n,L = 3,eta0_multiplier = 1000,lambda_grid = 0,
                 min_abs_corr = 0,max_iter = 100,verbose = TRUE)
out$sets <- list(cs = list(CS1 = get_cs(out$selected_fit$alpha[1,]),
                           CS2 = get_cs(out$selected_fit$alpha[2,]),
                           CS3 = get_cs(out$selected_fit$alpha[3,])))
pip_plot(out,causal_snps)
```

On the other hand, when the prior is largely uninformative, Dat's
method is very conservative, and only recovers one of the two causal SNPs:

```{r aiw-2, fig.height=1.75, fig.width=5.25, results="hide", message=FALSE}
out <- fit_aiwn(v,R_out,n,L = 3,eta0_multiplier = 1,lambda_grid = 0,
                min_abs_corr = 0,max_iter = 100,verbose = TRUE)
out$sets <- list()
out$sets <- list(cs = list(CS1 = get_cs(out$selected_fit$alpha[1,])))
pip_plot(out,causal_snps)
```

Ideally there should be a "middle ground" that finds a good tradeoff
between identifying both CSs and recognizing the inaccuracies in the
LD, but I was unsuccessful:

```{r aiw-3, fig.height=1.75, fig.width=5.25, results="hide", message=FALSE}
out <- fit_aiwn(v,R_out,n,L = 3,eta0_multiplier = 10,lambda_grid = 0,
                 min_abs_corr = 0,max_iter = 100,verbose = TRUE)
out$sets <- list(cs = list(CS1 = get_cs(out$selected_fit$alpha[1,]),
                           CS2 = get_cs(out$selected_fit$alpha[2,]),
                           CS3 = get_cs(out$selected_fit$alpha[3,])))
pip_plot(out,causal_snps)
```

Next, I'm trying out the "collapsed" method developed by Dat. Although
it has some disappointments, it seems to be achieving closer to the
behaviour that we are looking for, at least in this one example.

When the prior is very informative (large $\eta_0$), the result is
very similiar to `susie_rss()` (which does not model LD mismatch):

```{r collapsed-1, fig.height=1.75, fig.width=5.25}
source("code/01-collapsed-R.R")
source("code/03-SER-fallback-and-plot.R")
out1 <- susie_iw(v,R_out,n,L = 3,lambda_grid = 0,eta0 = 1000,
                max_iter = 100,tol = 1e-6,min_abs_corr = 0,
				verbose = FALSE)
pip_plot(out1,causal_snps)
```

On the other hand, when the prior is largely uninformative (small
$\eta_0$), the method is conservative, and recovers only one of the
two causal SNPs:

```{r collapsed-2, fig.height=1.75, fig.width=5.25}
out2 <- susie_iw(v,R_out,n,L = 3,lambda_grid = 0,eta0 = 75,
                 max_iter = 100,tol = 1e-6,min_abs_corr = 0,
		 		 verbose = FALSE)
pip_plot(out2,causal_snps)
```

The collapsed method is able to achieve a "middle ground" (here I set
$\eta_0 = 200$) that identifies both causal SNPs without also
returning a third (false positive) causal SNP:

```{r collapsed-3, fig.height=1.75, fig.width=5.25}
out3 <- susie_iw(v,R_out,n,L = 3,lambda_grid = 0,eta0 = 125,
                 max_iter = 100,tol = 1e-6,min_abs_corr = 0,
		 		 verbose = FALSE)
pip_plot(out3,causal_snps)
```

The main disappointment with this result is that in achieving this
middle ground, the method has also become more uncertain about which
SNP is the causal SNP within each CS. (Although that may not
necessarily be a bad thing—*to discuss*.)

Another thing is that the ELBO from the different fits does not select
the "right" result:

```{r compare-collapsed-models}
max(out1$elbo)
max(out2$elbo)
max(out3$elbo)
```

Finally, I'd like to explore a very simple idea that has some of the
features we are seeking but not others. This idea is not intended to
be a real method but just a way of illustrate some of the ideas we are
exploring. Two of the features it has that we are seeking are:

1. It has a parameterized LD matrix ${\bf R}$ that can be adapted to
   the data. (Either implicitly or explicitly, like it is here.)

2. Adapting R should focus only on the correlations with the candidate
   causal SNPs; due to the "irrelevance of null SNPs" property, these
   are the only correlations that matter.

```{r simple-idea, message=FALSE}
nw <- 20
dat <- data.frame(weight = seq(0,0.005,length.out = nw),V = 0)
for (i in 1:nw) {
  R_mix <- cov2cor(R_out + dat[i,"weight"]*tcrossprod(z))
  fit_mix <- susie_rss(zhat,R_mix,n = n,L = 3,estimate_prior_method = "EM",
                       tol = 1e-8,min_abs_corr = 0,verbose = FALSE)
  dat[i,"V"] <- min(fit_mix$V)
}
```

This simple method has a single parameter (a weight) that interpolates
between the out-of-sample LD matrix and the cross-product of the
z-scores. So we can vary this weight to produce different fine-mapping
results. One problem is that we don't have a good way of selecting the
weight, but it is perhaps interesting that there is a "sweet spot" of
weights in which `susie_rss()` shrinks the third CS close to zero:

```{r simple-idea-2, fig.height=2.75, fig.width=3}
par(mar = c(4,4,2,1))
plot(dat$weight,dat$V,pch = 20,xlab = "weight")
lines(dat$weight,dat$V)
```

Indeed, for a carefully chosen weight, the result looks very close to
what we got from using the in-sample LD:

```{r simple-idea-pip-plot, fig.height=1.75, fig.width=5.25, message=FALSE}
w <- dat[which.min(dat$V),"weight"]
R_mix <- cov2cor(R_out + w*tcrossprod(z))
fit_mix <- susie_rss(zhat,R_mix,n = n,L = 3,estimate_prior_method = "EM",
                     tol = 1e-8,min_abs_corr = 0,verbose = FALSE)
pip_plot(fit_mix,causal_snps)
```

To be clear this is just intended to explore the ideas—it isn't practical.
